---
title: "RSN_Data_exploration"
author: "OLH"
date: "2026-02-09"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```

# Summary of Q's


# Downlaod data
```{r}

# create data folders
dir.create(file.path("2026 Data"), showWarnings = F)
dir.create(file.path("2026 Data", 'RSN_Site'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data', 'Updated'), showWarnings = F)

# Identify desired files to download
site_info_file <- "RSN_Master_Site.csv"
qaqc_data <- "RSN_Inventory_QAQC_Final_2026-03-24.csv"

# Download data files
# site info
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1PWk45RgTK47B-PX71QG6BazpJyAtidQ1")) %>% 
  dplyr::filter(name == site_info_file) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "RSN_Site", .$name))

# qaqc data
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1VR7WRVcF3behOWDoXtuFlN9wK0M24jOb")) %>% 
  dplyr::filter(name == qaqc_data) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "QAQC_Data", "Updated", .$name))

# Read in site info and data files
site_info <- read.csv(file = file.path("2026 Data", 'RSN_Site', 'site_metadata_OLH.csv'))

data <-  read.csv(file = file.path("2026 Data", 'QAQC_Data','Updated', qaqc_data))

```


# Length and No of measurements
```{r}
unique(data$SITE)
data_summary <- data %>%
  # make a column for whether or not tree is above or below 2.5 cm
  mutate(over2.5 = ifelse(DBH>2.5, 1, 0)) %>%
  # make a column that indicates, dead, live, or seedling less than 2.5 cm
  mutate(tree_status = case_when(
    over2.5 == 1 & DA == 1 ~ as.character("Live"), 
    over2.5 == 0 & DA ==1 ~ as.character("Seedling"),
    DA == 0 ~ as.character("Dead"),
    DBH_raw == -8888 ~ as.character("Missing"), 
    DBH_raw %in% c(-9999, -7777) ~ as.character("Dead"),
    TRUE ~ "Missing"
  )) %>%
  group_by(SITE, Year, tree_status) %>%
  summarize(no_plots = n_distinct(PLOT), no_trees = n_distinct(TREE)) %>%
  pivot_wider(names_from = tree_status, values_from = c(no_plots, no_trees), values_fill = 0) %>%
  ungroup()%>%
  group_by(SITE) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(live_diff = no_trees_Live - lag(no_trees_Live))

# write.csv(data_summary,'/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_year_tree_summary.csv', row.names = FALSE)


# 1. Open the PDF file
# pdf("/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/Site_Trends.pdf", width = 10, height = 7)

# 2. Get a list of all sites
plot_ready <- data_summary %>%
  # Select only the columns needed for the graph
  select(SITE, Year, no_trees_Live, no_trees_Dead, no_trees_Seedling) %>%
  # Pivot so 'Live' and 'Dead' are in one column
  pivot_longer(cols = c(no_trees_Live, no_trees_Dead, no_trees_Seedling), 
               names_to = "Status", 
               values_to = "Count") %>%
  # Clean up labels: "no_trees_Live" becomes "Live"
  mutate(Status = str_remove(Status, "no_trees_"))
site_list <- unique(plot_ready$SITE)

for (s in site_list) {
  p <- plot_ready %>%
    filter(SITE == s) %>%
    ggplot(aes(x = Year, y = Count, color = Status, group = Status)) +
    geom_line(size = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("Live" = "#2ecc71", "Dead" = "#e74c3c", "Seedling" = "skyblue")) +
    labs(title = paste("Tree Density Trends - Site:", s),
         subtitle = "Comparing Live vs. Dead stems over time",
         x = "Year", y = "Number of Trees") +
    theme_minimal()
  
  print(p) # This "prints" the plot as a new page in the PDF
}

```

# Create a table for the years and sites

```{r}
str(data_summary)
site_yrs <- data %>%
  select(SITE, Year, PLOT) %>%
  distinct() %>%
  group_by(SITE, Year) %>%
  summarize(n_plot = n_distinct(PLOT)) %>%
  arrange(Year) %>%
  mutate(present = 1) %>%
  left_join(site_info, by = "SITE")

# # divide sites
# older <- site_yrs %>%
#   filter(Year < 2013)
unique(site_yrs$age_class)
mature <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Mature"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Mature Stands")
mature

inter <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Intermediate"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Intermediate Stands")
inter

young <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Young"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Young Stands")
young

# unk <- ggplot(data = subset(site_yrs, is.na(site_yrs$age_class)), aes(SITE, Year)) + 
#   geom_point() +
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
#   ggtitle("Unknown Stands")
# unk

unique(data$Year)



```

# No plots (v1)
- this was done on older data...

```{r}
# unique(data$SITE)
# data_yrs <- data %>%
#   # remove any trailing spaces
#   mutate(SITE = str_trim(SITE, side = "right"))%>%
#   select(SITE, PLOT, Year) %>%
#   distinct() 
# 
# ggplot(data_yrs, aes(SITE, Year))+
#   geom_point()+
#   theme_minimal()+
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
# 
# # how many plots in each year
# no_plots <- data %>%
#   # remove any trailing spaces
#   mutate(SITE = str_trim(SITE, side = "right"))%>%
#   select(SITE, PLOT, Year) %>%
#   distinct() %>%
#   group_by(SITE, Year) %>%
#   summarize(no_plots = n_distinct(PLOT)) %>%
#   rename(sitename = SITE) 
# 
# no_plots.filter <- no_plots %>%
#   group_by(sitename) %>%
#   filter(all(no_plots %in% c(12, 30))) %>%
#   ungroup() %>%
#   select(sitename, no_plots) %>%
#   distinct()
# 
# no_plots.filter.opp <- no_plots %>%
#   filter(sitename %!in% no_plots.filter$sitename) %>% 
#   select(sitename) %>%
#   distinct()
  
# write.csv(no_plots, '/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/no_plots.csv')
```
## Histograms of dbh
```{r}
data.dbh <- data %>%
  filter(!is.na(DBH))
ggplot(data.dbh, aes(DBH)) +
  geom_histogram()+
  theme_bw()+
  facet_wrap(~SPECIES, scales="free")
  
salbeb <- data %>%
  filter(SPECIES == "SALBEB")
  

```

## Data Prep
- remove "SALBEB" - only one instance where it was tagged
- remove FP1A - this site is a bit of a mess in terms of long term data
```{r}
data.2 <- data %>%
  # remove salbeb
  filter(SPECIES != "SALBEB") %>%
  # remove site FP1A
  filter(SITE != "FP1A")

```

# Missing Trees?
## Update Notes - April - wiht udpated data from Zach
- 120 observation where a tree exceeds 2.5 and then dips back down
- there are 10 trees that are only ever NA,and seems that first observation is always something along line of dbh not recorded, but it is marked as alive
- when get rid of the trees that are below 2.5cm DBH, there are 1721 trees that are only NA (indicating likely that they died) and this amounts to 2715 NA observations - will want to remove these from the final datasets
- 785 trees that seem to go missing; last measurement does not indicate that they are dead; but importantly this goes to 480 when excluding sites that burned during the record (so close to half go missing b/c they burned). Further, UP2A accounts for a lot of the missing trees, and there was a change in sampling effort in time. So excluding that site as well, it is about 250 trees that go missing. A few sites have around 20 in a year, but by and large, it is pretty low.
- 906 trees that are dead and go missing without having ever being recorded as DOWN - goes down to 761 if excluding the burned sites and the UP2A with the weird samplign effort stuff. IMportantly all of those that fit into this category are from the UP or FP sites and by in large, in the earlier sampling periods. Likely reflects changes in methodology wiht the official establishment of the RSN network. 
- now only 23 that seem to maybe come back to life and represent new growht
- there are 9 that have the first observaiton listed as DA - 6 of these were ones showing new growth (i.e., "measured standing dead"), two are missing DBH (NA), and one has a note that marked 

## Notes from March
- 1967 trees with NA values only ever - mostly the standing dead that were measured
- There are 859 trees that end up going missing that were measured and their last measurement does not indicate that they are dead (837 if update to 2.5 vs. 2.4 AND make a few changes on teh tree number (see section above)) AND actually if update so that trees shrinking are included, this drops to 796
- 47 site-yeras with more than 1 tree missing, and only 19 with more than 5. going to look into this more. goes down to 17 with the updated code allowing for trees to shrink
- want to make sure that any tree that is going to be recorded as dead had a record before...?
- there are 3116 observations of only NA, but its only 1972 distinct trees - a lot of these are because they had DBH less than 2.5 in iniital obs, because if I remove that requirement, then there are 154 obs corresponding to 85 trees - this doesn't really affect the last measurement, but will be important to remove (?) for demographic surveys
- there are also a number of trees that are dead and not tracked through time but still seem to be standing - 1443 trees and most seem to be from BDI5 and BDI6 (makes sense, they burned) and then floodplain sites and upland - so there does some to be some loss of the tracking over time - all stsanding dead or dead DBH measured; also about 500 of these were the initially measured standing dead notes
- seems that there are 121 observations where a tree exceed 2.5 DBH and dipped back down
- there are 11,713 observations that are only ever dead. most of them were measured in that initial campaign of 2013-2014 where the trees measured as standing dead
- from the dataset that then cleans all of this there are 162 instances of a tree showing new growth, as in it was recorded dead (DA ==0) but then later came to be recorded as alive with DBH which account for 383 obs across in teh dataset. majority of these observations are in 2018 and forward. A lot of these come from the standing dead measurement (only 14 trees do not originate there)
```{r}
# get rid of the small trees
data.25 <- data.2 %>%
  filter(SPECIES != "SALBEB") %>%
  arrange(UniqueID, Year) %>% # Ensure data is in chronological order
  group_by(UniqueID) %>%
  # keep NA DBH, DBH > 2.5, or if the DBH exceeds 2.5, it will keep subsequent observatoins
  filter(is.na(DBH) | DBH >= 2.49 | cumany(DBH >= 2.49)) %>%
  ungroup()

# this code above removes 4693 observations

# check to see which trees/how many trees dip below 2.5 DBH over time after measuring above
data.252 <- data.2 %>%
  filter(is.na(DBH) | DBH > 2.49) 

anti <- anti_join(data.25, data.252)

# see which trees only have an NA - this is after the seedlings have been removed, so in many cases these had an intiial measurement that was NA and corresponded to somehting like 0.7cm
na.trees <- data.25 %>%
  group_by(UniqueID) %>%
  filter(all(is.na(DBH))) #%>%
  #distinct(UniqueID)

# remove the na trees only
data.25 <- data.25 %>%
  filter(UniqueID %!in% na.trees$UniqueID)

# Derive site visits from tree data if no separate visit table
site_visits <- data.25 %>%
  distinct(SITE, Year)

#write.csv(site_visits,'/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_visits.csv', row.names = FALSE)
#Find the last occasion each tree was recorded ---
last_seen <- data.25 %>%
  group_by(UniqueID, SITE) %>%
  summarise(last_occasion = max(Year), .groups = "drop")

# Get all site visits that occurred AFTER each tree's last record 
# Cross join trees to visits, then filter to later visits
later_visits <- last_seen %>%
  left_join(site_visits, by = "SITE") %>%       # all visits for that site
  filter(Year > last_occasion) %>%                  # only visits after last record
  # mutate to create a new column that has the difference between teh year and last occassion adn then for each site/tree, select the smallest value
  mutate(diff = Year - last_occasion) %>%
  group_by(UniqueID, SITE) %>%
  slice_min(diff, n=1)
  
# select the trees identified in the data file, and filter that those observations are not dead
data.select <- data.25 %>%
  filter(UniqueID %in% later_visits$UniqueID) %>%
  group_by(UniqueID, SITE) %>%
  slice_max(Year, n =1) %>%
  filter(DA == 1)

# filter out the sites that burned
burned <- c("BDI5", "BDI6", "FP5C", "FP1C")

data.select.burn <- data.select %>%
  filter(SITE %!in% burned)

# summarize this by site
site_miss_tree <- data.select.burn %>%
  ungroup() %>%
  group_by(SITE, Year) %>%
  dplyr::count() %>%
  filter(n>5)

# select trees identified as missing that are dead but not yet fallen
data.select.2 <- data.25 %>%
  filter(UniqueID %in% later_visits$UniqueID) %>%
  group_by(UniqueID, SITE) %>%
  slice_max(Year, n =1) %>%
  filter(DA == 0)%>%
  filter(DOWN == 1) %>%
  filter(SITE %!in% burned) %>%
  filter(SITE != "UP2A")

# summarize this at the site level
site_miss_tree.2 <- data.select.2 %>%
  ungroup() %>%
  group_by(SITE, Year) %>%
  dplyr::count() %>%
  filter(n>5)

# Which trees only ever register as dead - almost all are likely from teh campgain that initially measured the standing dead - Zach has dealt with this in teh code!

dead <- data.25 %>%
  group_by(UniqueID) %>%
  filter(all(DA == 0))
stdm <- data.25 %>%
  filter(OfficeNotes == "Trees Measured Standing dead as a part of initial site measurements in 2013 and 2014. Methodology and measurements should not have continued for this tree in subsequent years.")

stdmt <- data.25 %>%
  filter(UniqueID %in% stdm$UniqueID)

diff <- anti_join(stdmt, dead)

# will get rid of trees that are only dead
data.demo <- data.25 %>%
  filter(UniqueID %!in% dead$UniqueID)
  
# how many show the first observation dead and then come back to life
dead_to_life <- data.demo %>%
  arrange(UniqueID, Year) %>%
  group_by(UniqueID) %>%
  # Create a flag: Was this tree ever recorded as dead in a PREVIOUS year?
  mutate(was_dead = lag(cumany(DA == 0), default = FALSE)) %>%
  # Filter for rows where it is currently alive but was previously dead
  filter(DA == 1 & was_dead) %>%
  ungroup()

dtlt <- data.demo %>%
  filter(UniqueID %in% dead_to_life$UniqueID) %>%
  # try then getting rid of the ones that resulted from the standing dead measuresmetns
  filter(UniqueID %in% stdm$UniqueID) 

# how many have the first observation that is dead
first.dead <- data.demo %>%
  # Ensure the data is in chronological order for each tree
  group_by(UniqueID) %>%
  arrange(Year, .by_group = TRUE) %>%
  
  # Keep only the first record for each tree
  slice(1) %>%
  
  # Filter for those that were dead (DA == 0)
  filter(DA == 0) %>%
  ungroup()
```

# Join Data wiht metadata
- here we will use data.25 for following analyses, this removes seedlings BUT importantly keeps in trees that shrink after an observation exceeds 2.5 cm. This then removes trees that are only ever recorded as NA
- not data.demo because none that are only dead in this!
```{r}
# read in the site-year data
site_year <- read.csv(file = file.path("2026 Data", 'Data_checks', 'site_years.csv'))

# join the site-year data with the overall site-metadata
site_data <- left_join(site_year, site_info, by = "SITE")

data.25.site <- left_join(data.25, site_data, by = c("SITE", "Year"))

# graph of sites and age classes
site.yr.sum <- site_year %>%
  group_by(age_class_yr, sample.rd) %>%
  summarize(num=n())

ggplot(site.yr.sum, aes(sample.rd, num, group=age_class_yr, fill=age_class_yr))+
  geom_col(position="dodge")+
  theme_bw()

```

# DBH THROUGH TIME
- tree level
- approximately 29ish have a DA of 0 that have a DBH; DBH is not NA; some of these are the standing dead measured; should probably dive into this and figure out a few more. Could consider just getting rid of the dead, but going to keep for now. 
```{r}
# try getting rid of the NAs
data.dbh <- data.25 %>%
  filter(!is.na(DBH)) %>%
  group_by(UniqueID) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    DBH_rel2 = pi * (DBH/2)^2,
    BA = 0.00007854 * (DBH)^2,
    DBH_diff = DBH - lag(DBH),
    DBH_rel_diff = ((DBH_rel2 - lag(DBH_rel2))/lag(DBH_rel2))*100,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = DBH_diff / Year_diff,
    DBH_relativized_BA = DBH_rel_diff/Year_diff,
    BA_growth = (BA_diff/Year_diff)  * 10000,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    RBA = (log(BA) - log(lag(BA)))/Year_diff,
    Year_before = lag(Year)
  ) %>%
  #filter(!is.na(Annual_Growth)) %>%
  left_join(site_data, by = c("SITE", "Year"))

data.dbh$age_class_yr <- factor(data.dbh$age_class_yr,
                   levels = c("Young", "Intermediate", "Mature"))

```

## Some initial viz

### DBH by age and species

```{r}
overall.dbh <- ggplot(data = data.dbh, aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr)+
  ggtitle("All Trees")
overall.dbh
  
picmar.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "PICMAR"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("PICMAR")
picmar.dbh

picgla.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "PICGLA"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("PICGLA")
picgla.dbh

betneo.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "BETNEO"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("BETNEO")
betneo.dbh

poptre.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "POPTRE"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("POPTRE")
poptre.dbh

popbal.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "POPBAL"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("POPBAL")
popbal.dbh

larlar.dbh <- ggplot(data = subset(data.dbh, data.dbh$SPECIES == "LARLAR"), aes(DBH))+
  geom_histogram()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")+
  ggtitle("LARLAR")
larlar.dbh
  
```



### PicMAR
```{r, fig.width=12, fig.height=6, fig.retina=2}
picmar <- data.dbh %>%
  filter(SPECIES == "PICMAR")
picmar.rgr <- ggplot(data = picmar, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.rgr

picmar.rba <- ggplot(data = subset(picmar, picmar$DBH_relativized_BA < 100), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.rba

picmar.ba <- ggplot(data = picmar, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.ba

picmar.ann <- ggplot(data = picmar, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.ann
```

### PICGLA

```{r, fig.width=12, fig.height=6, fig.retina=2}

picgla <- data.dbh %>%
  filter(SPECIES == "PICGLA")
picgla.rgr <- ggplot(data = picgla, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.rgr

picgla.rba <- ggplot(data = subset(picgla, picgla$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.rba

picgla.ba <- ggplot(data = picgla, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.ba

picgla.ann <- ggplot(data = picgla, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.ann

```

### POPTRE

```{r, fig.width=12, fig.height=6, fig.retina=2}

poptre <- data.dbh %>%
  filter(SPECIES == "POPTRE")
poptre.rgr <- ggplot(data = poptre, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.rgr

poptre.rba <- ggplot(data = poptre, aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.rba

poptre.ba <- ggplot(data = poptre, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.ba

poptre.ann <- ggplot(data = poptre, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.ann

```
### BETNEO
```{r, fig.width=12, fig.height=6, fig.retina=2}

betneo <- data.dbh %>%
  filter(SPECIES == "BETNEO")
betneo.rgr <- ggplot(data = betneo, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.rgr

betneo.rba <- ggplot(data = subset(betneo, betneo$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.rba

betneo.ba <- ggplot(data = betneo, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.ba

betneo.ann <- ggplot(data = betneo, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.ann

```

### POPBAL

```{r, fig.width=12, fig.height=6, fig.retina=2}

popbal <- data.dbh %>%
  filter(SPECIES == "POPBAL")
popbal.rgr <- ggplot(data = popbal, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.rgr

popbal.rba <- ggplot(data = subset(popbal, popbal$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.rba

popbal.ba <- ggplot(data = popbal, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.ba

popbal.ann <- ggplot(data = popbal, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.ann

```


##LARLAR

```{r, fig.width=12, fig.height=6, fig.retina=2}

larlar <- data.dbh %>%
  filter(SPECIES == "LARLAR")
larlar.rgr <- ggplot(data = larlar, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.rgr

larlar.rba <- ggplot(data = subset(larlar, larlar$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.rba

larlar.ba <- ggplot(data = larlar, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.ba

larlar.ann <- ggplot(data = larlar, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.ann

```

# Site level tree growth
- going to summarize by site and then by sample.yr, by tree species and age class
- had initially done a summary by decade adn year, but I have udpated the code to have a sample.rd that helps to combine sites that have multiple observations

```{r}
site_growth <- data.dbh %>%
  ungroup() %>%
  filter(!is.na(DBH_diff)) %>%
  # get rid of outliers 
  filter(Annual_Growth < 4 & Annual_Growth > -4) %>%
  group_by(sample.rd, sample.yr, SPECIES, age_class_yr, SITE) %>%
  summarize(n_tree = n_distinct(UniqueID), n_plot = n_distinct(PLOT),
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()),
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_BG = mean(BA_growth), median_BG = median(BA_growth), se_BG = sd(BA_growth)/sqrt(n()),
            mean_RBAG = mean(DBH_relativized_BA), median_RBAG = median(DBH_relativized_BA), se_RBAG = sd(DBH_relativized_BA)/sqrt(n()),
            mean_RBA = mean(RBA), median_RBA = median(RBA), se_RBA = sd(RBA)/sqrt(n())
            ) %>%
  arrange(SPECIES, sample.rd, sample.yr, age_class_yr) %>%
  # remove tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR")) 
  # get rid of obs with only 1 site
  # filter(n_plot >1) %>%
  # # requrie at least 5 trees
  # filter(n_tree >5)


site_growth_ts <- data.dbh %>%
  ungroup() %>%
  # make a new column for if a tree is large or small
  mutate(size = ifelse(DBH>10, "Large", "Small")) %>%
  filter(!is.na(DBH_diff)) %>%
  # Fitler out outliers in growht
  filter(Annual_Growth < 4 & Annual_Growth > -4) %>%
  group_by(sample.rd,sample.yr, SPECIES, age_class_yr, SITE, size) %>%
  summarize(n_tree = n_distinct(UniqueID), n_plot = n_distinct(PLOT),
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()),
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_BG = mean(BA_growth), median_BG = median(BA_growth), se_BG = sd(BA_growth)/sqrt(n()),
            mean_RBAG = mean(DBH_relativized_BA), median_RBAG = median(DBH_relativized_BA), se_RBAG = sd(DBH_relativized_BA)/sqrt(n()),
            mean_RBA = mean(RBA), median_RBA = median(RBA), se_RBA = sd(RBA)/sqrt(n())
            ) %>%
  arrange(SPECIES, sample.rd, sample.yr, age_class_yr, size) %>%
# remove tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR"))
  # get rid of obs with only 1 site
  # filter(n_plot >1) %>%
  # # requrie at least more than 4 trees (so min 5)
  # filter(n_tree >4)

```

# Visualize growth over time

## Across sizes
```{r}
ggplot(site_growth, aes(sample.yr, n_tree, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(site_growth, aes(sample.yr, n_plot, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(site_growth, aes(sample.yr, mean_dbh, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh - se_dbh, ymax = mean_dbh + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(site_growth, aes(sample.yr, mean_AG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG - se_AG, ymax = mean_AG + se_AG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(site_growth, aes(sample.yr, mean_RGR, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR - se_RGR, ymax = mean_RGR + se_RGR),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(site_growth, aes(sample.yr, mean_BG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG - se_BG, ymax = mean_BG + se_BG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(site_growth, aes(sample.yr, mean_RBAG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG - se_RBAG, ymax = mean_RBAG + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(site_growth, aes(sample.yr, mean_RBA, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBA - se_RBA, ymax = mean_RBA + se_RBA),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")
```

## By size
```{r}
ggplot(site_growth_ts, aes(sample.yr, n_tree, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(site_growth_ts, aes(sample.yr, n_plot, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(site_growth_ts, aes(sample.yr, mean_dbh, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh - se_dbh, ymax = mean_dbh + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(site_growth_ts, aes(sample.yr, mean_AG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG - se_AG, ymax = mean_AG + se_AG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")


ggplot(site_growth_ts, aes(sample.yr, mean_RGR, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR - se_RGR, ymax = mean_RGR + se_RGR),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")


ggplot(site_growth_ts, aes(sample.yr, mean_BG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG - se_BG, ymax = mean_BG + se_BG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")


ggplot(site_growth_ts, aes(sample.yr, mean_RBAG, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG - se_RBAG, ymax = mean_RBAG + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(site_growth_ts, aes(sample.yr, mean_RBA, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 1.3), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBA - se_RBA, ymax = mean_RBA + se_RBA),
    width = 0.2,
    position = position_dodge(width = 1.3)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")
```
# Summarize growth across sites

```{r}
ac_sample <-  site_growth%>%
  ungroup()%>%
  group_by(sample.yr, SPECIES, age_class_yr) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_BG2 = mean(mean_BG), se_BG = sd(mean_BG)/sqrt(n()),
            mean_RBAG2 = mean(mean_RBAG), se_RBAG = sd(mean_RBAG)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            ) #%>%
  # must be in more than one site
  #filter(n_site > 1)

ac_sample.size <- site_growth_ts %>%
    ungroup()%>%
  group_by(sample.yr, SPECIES, age_class_yr, size) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_BG2 = mean(mean_BG), se_BG = sd(mean_BG)/sqrt(n()),
            mean_RBAG2 = mean(mean_RBAG), se_RBAG = sd(mean_RBAG)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            )  #%>%
  #filter(n_site > 1)
```

## Visuals for aggregated

### Age Class
```{r}
ggplot(ac_sample, aes(sample.yr, n_tree2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample, aes(sample.yr, n_site, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample, aes(sample.yr, mean_dbh2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh2 - se_dbh, ymax = mean_dbh2 + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample, aes(sample.yr, mean_AG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG2 - se_AG, ymax = mean_AG2 + se_AG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(ac_sample, aes(sample.yr, mean_RGR2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR2 - se_RGR, ymax = mean_RGR2 + se_RGR),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample, aes(sample.yr, mean_BG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG2 - se_BG, ymax = mean_BG2 + se_BG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(ac_sample, aes(sample.yr, mean_RBAG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG2 - se_RBAG, ymax = mean_RBAG2 + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample, aes(sample.yr, mean_RBA2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBA2 - se_RBA, ymax = mean_RBA2 + se_RBA),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

```


### AgeClass-Size
```{r}
ggplot(ac_sample.size, aes(sample.yr, n_tree2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(ac_sample.size, aes(sample.yr, n_site, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(ac_sample.size, aes(sample.yr, mean_dbh2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh2 - se_dbh, ymax = mean_dbh2 + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_sample.size, aes(sample.yr, mean_AG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG2 - se_AG, ymax = mean_AG2 + se_AG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")


ggplot(ac_sample.size, aes(sample.yr, mean_RGR2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR2 - se_RGR, ymax = mean_RGR2 + se_RGR),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(ac_sample.size, aes(sample.yr, mean_BG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG2 - se_BG, ymax = mean_BG2 + se_BG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")


ggplot(ac_sample.size, aes(sample.yr, mean_RBAG2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG2 - se_RBAG, ymax = mean_RBAG2 + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")

ggplot(ac_sample.size, aes(sample.yr, mean_RBA2, group = age_class_yr, color=age_class_yr))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBA2 - se_RBA, ymax = mean_RBA2 + se_RBA),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES+size, nrow=3, scales="free")
```

## Just by size not divided by age class

```{r}
ac_size <- site_growth_ts %>%
    ungroup()%>%
  group_by(sample.yr, SPECIES, size) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_BG2 = mean(mean_BG), se_BG = sd(mean_BG)/sqrt(n()),
            mean_RBAG2 = mean(mean_RBAG), se_RBAG = sd(mean_RBAG)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            )

ggplot(ac_size, aes(sample.yr, n_tree2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_size, aes(sample.yr, n_site, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_size, aes(sample.yr, mean_dbh2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh2 - se_dbh, ymax = mean_dbh2 + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_size, aes(sample.yr, mean_AG2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG2 - se_AG, ymax = mean_AG2 + se_AG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(ac_size, aes(sample.yr, mean_RGR2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR2 - se_RGR, ymax = mean_RGR2 + se_RGR),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_size, aes(sample.yr, mean_BG2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG2 - se_BG, ymax = mean_BG2 + se_BG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(ac_size, aes(sample.yr, mean_RBAG2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG2 - se_RBAG, ymax = mean_RBAG2 + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(ac_size, aes(sample.yr, mean_RBA2, group = size, color=size))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBA2 - se_RBA, ymax = mean_RBA2 + se_RBA),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")
```

# Plot and Site level details...?
- working with data.demo now - this data has removed trees that only ever have a DBH of NA as well as ones that are always dead (primarily resulting from trees where initial surveys measured standing dead) and DBH had to be over 2.49, but it allows for shrinkage if a tree exceeds that threshold
- also includes the initial data filtering of weird tree numbers, etc!

## Plot basal area over time

## Density 

```{r}
plot_data <- data.25 %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot.Area.M2) %>%
  summarize(n=n(), num_alive = sum(DA), num_dead = n-num_alive) %>%
  mutate(prop_alive = (num_alive/n)*100) %>%
  left_join(site_year, by =c("SITE", "Year"))


# make it wider to look at full plot level dynamics
plot_wide <- plot_data %>%
  pivot_wider(names_from = SPECIES,
              values_from=c("n", "num_alive", "prop_alive"), values_fill=0) %>%
  mutate(tot_trees = n_POPBAL + n_POPTRE + n_PICGLA + n_PICEA + n_PICMAR + n_LARLAR + n_BETNEO,
         tot_alive = num_alive_POPBAL + num_alive_POPTRE + num_alive_PICGLA + num_alive_PICEA + num_alive_PICMAR + num_alive_LARLAR + num_alive_BETNEO,
         prop_alive = tot_alive/tot_trees)

# scale to the site level
plot_wide$area.samp <- plot_wide$Plot.Area.M2 * plot_wide$sampled_plots

# summarize at site level 
str(plot_data)
plot_data.2 <- plot_data %>%
  ungroup()%>%
  group_by(sample.yr, SITE, SPECIES, sampled_plots_period, Plot.Area.M2, age_class_yr) %>%
  summarize(no_trees = sum(n), no_alive = sum(num_alive), sample = mean(sampled_plots), plot_area= mean(Plot.Area.M2), no_trees_pp = no_trees/sample, no_alive_pp = no_alive/sample, prop_alive_pp = no_alive_pp/no_trees_pp, no_trees_m2 = no_trees_pp/plot_area, no_alive_m2 = no_alive_pp/plot_area, prop_alive_m2 = no_alive_m2/no_trees_m2)

plot_wide.2 <- plot_data.2 %>%
  select(-c(no_trees, no_alive, sample, no_trees_pp, no_alive_pp, prop_alive_pp))%>%
  ungroup()%>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("no_trees_m2", "no_alive_m2", "prop_alive_m2"), values_fill=0) %>%
  mutate(tot_trees = no_trees_m2_POPBAL + no_trees_m2_POPTRE + no_trees_m2_PICGLA + no_trees_m2_PICEA + no_trees_m2_PICMAR + no_trees_m2_LARLAR + no_trees_m2_BETNEO, 
         tot_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_PICGLA + no_alive_m2_PICEA + no_alive_m2_PICMAR + no_alive_m2_LARLAR + no_alive_m2_BETNEO, 
         prop_alive = tot_alive/tot_trees, decid_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_BETNEO, con_alive = no_alive_m2_PICGLA + no_alive_m2_PICEA + no_alive_m2_PICMAR + no_alive_m2_LARLAR) 


# get the units right!

ggplot(plot_wide.2, aes(sample.yr, tot_trees, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide.2, aes(sample.yr, tot_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Alive Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide.2, aes(sample.yr, decid_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Deciduous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide.2, aes(sample.yr, con_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Coniferous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide.2, aes(sample.yr, prop_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Proportion Alive")+
  facet_wrap(~age_class_yr)
```

## Density of large trees

```{r}
plot_large <- data.25 %>%
  filter(DBH >=10) %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot.Area.M2) %>%
  summarize(n=n(), num_alive = sum(DA), num_dead = n-num_alive) %>%
  mutate(prop_alive = (num_alive/n)*100) %>%
  left_join(site_year, by =c("SITE", "Year"))


# make it wider to look at full plot level dynamics
plot_wide_large <- plot_large %>%
  pivot_wider(names_from = SPECIES,
              values_from=c("n", "num_alive", "prop_alive"), values_fill=0) %>%
  mutate(tot_trees = n_POPBAL + n_POPTRE + n_PICGLA + n_PICMAR + n_LARLAR + n_BETNEO,
         tot_alive = num_alive_POPBAL + num_alive_POPTRE + num_alive_PICGLA  + num_alive_PICMAR + num_alive_LARLAR + num_alive_BETNEO,
         prop_alive = tot_alive/tot_trees)

# scale to the site level
plot_wide_large$area.samp <- plot_wide_large$Plot.Area.M2 * plot_wide_large$sampled_plots

# summarize at site level 
plot_data.large<- plot_large %>%
  ungroup()%>%
  group_by(sample.yr, SITE, SPECIES, sampled_plots_period, Plot.Area.M2, age_class_yr) %>%
  summarize(no_trees = sum(n), no_alive = sum(num_alive), sample = mean(sampled_plots), plot_area= mean(Plot.Area.M2), no_trees_pp = no_trees/sample, no_alive_pp = no_alive/sample, prop_alive_pp = no_alive_pp/no_trees_pp, no_trees_m2 = no_trees_pp/plot_area, no_alive_m2 = no_alive_pp/plot_area, prop_alive_m2 = no_alive_m2/no_trees_m2)

plot_wide_large <- plot_data.large %>%
  select(-c(no_trees, no_alive, sample, no_trees_pp, no_alive_pp, prop_alive_pp))%>%
  ungroup()%>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("no_trees_m2", "no_alive_m2", "prop_alive_m2"), values_fill=0) %>%
  mutate(tot_trees = no_trees_m2_POPBAL + no_trees_m2_POPTRE + no_trees_m2_PICGLA  + no_trees_m2_PICMAR + no_trees_m2_LARLAR + no_trees_m2_BETNEO, 
         tot_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_PICGLA + no_alive_m2_PICMAR + no_alive_m2_LARLAR + no_alive_m2_BETNEO, 
         prop_alive = tot_alive/tot_trees, decid_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_BETNEO, con_alive = no_alive_m2_PICGLA +  no_alive_m2_PICMAR + no_alive_m2_LARLAR) 


# get the units right!
ggplot(plot_wide_large, aes(sample.yr, tot_trees, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_large, aes(sample.yr, tot_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Alive Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_large, aes(sample.yr, decid_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Deciduous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_large, aes(sample.yr, con_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Coniferous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_large, aes(sample.yr, prop_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Proportion Alive")+
  facet_wrap(~age_class_yr)
```


## Density of small trees

```{r}
plot_small <- data.25 %>%
  filter(DBH < 10) %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot.Area.M2) %>%
  summarize(n=n(), num_alive = sum(DA), num_dead = n-num_alive) %>%
  mutate(prop_alive = (num_alive/n)*100) %>%
  left_join(site_year, by =c("SITE", "Year"))


# make it wider to look at full plot level dynamics
plot_wide_small <- plot_small %>%
  pivot_wider(names_from = SPECIES,
              values_from=c("n", "num_alive", "prop_alive"), values_fill=0) %>%
  mutate(tot_trees = n_POPBAL + n_POPTRE + n_PICGLA + n_PICMAR + n_LARLAR + n_BETNEO,
         tot_alive = num_alive_POPBAL + num_alive_POPTRE + num_alive_PICGLA  + num_alive_PICMAR + num_alive_LARLAR + num_alive_BETNEO,
         prop_alive = tot_alive/tot_trees)

# scale to the site level
plot_wide_small$area.samp <- plot_wide_small$Plot.Area.M2 * plot_wide_small$sampled_plots

# summarize at site level 
plot_data.small<- plot_small %>%
  ungroup()%>%
  group_by(sample.yr, SITE, SPECIES, sampled_plots_period, Plot.Area.M2, age_class_yr) %>%
  summarize(no_trees = sum(n), no_alive = sum(num_alive), sample = mean(sampled_plots), plot_area= mean(Plot.Area.M2), no_trees_pp = no_trees/sample, no_alive_pp = no_alive/sample, prop_alive_pp = no_alive_pp/no_trees_pp, no_trees_m2 = no_trees_pp/plot_area, no_alive_m2 = no_alive_pp/plot_area, prop_alive_m2 = no_alive_m2/no_trees_m2)

plot_wide_small <- plot_data.small %>%
  select(-c(no_trees, no_alive, sample, no_trees_pp, no_alive_pp, prop_alive_pp))%>%
  ungroup()%>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("no_trees_m2", "no_alive_m2", "prop_alive_m2"), values_fill=0) %>%
  mutate(tot_trees = no_trees_m2_POPBAL + no_trees_m2_POPTRE + no_trees_m2_PICGLA  + no_trees_m2_PICMAR + no_trees_m2_LARLAR + no_trees_m2_BETNEO, 
         tot_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_PICGLA + no_alive_m2_PICMAR + no_alive_m2_LARLAR + no_alive_m2_BETNEO, 
         prop_alive = tot_alive/tot_trees, decid_alive = no_alive_m2_POPBAL + no_alive_m2_POPTRE + no_alive_m2_BETNEO, con_alive = no_alive_m2_PICGLA +  no_alive_m2_PICMAR + no_alive_m2_LARLAR) 


# get the units right!
ggplot(plot_wide_small, aes(sample.yr, tot_trees, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_small, aes(sample.yr, tot_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Total Alive Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_small, aes(sample.yr, decid_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Deciduous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_small, aes(sample.yr, con_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Alive Coniferous Trees (m2)")+
  facet_wrap(~age_class_yr)

ggplot(plot_wide_small, aes(sample.yr, prop_alive, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  ylab("Proportion Alive")+
  facet_wrap(~age_class_yr)
```

# Mortality
```{r}
site_mortality <- data.25 %>%
  ungroup() %>%
  arrange(UniqueID, Year) %>%
  group_by(UniqueID) %>%
  # We look for the first index (row number) where DA is 0
  # but only if that index is NOT 1.
  filter(which(DA == 0)[1] > 1) %>%
  # Now that we've narrowed it to the right trees, 
  # we isolate that specific first-death row.
  slice(which(DA == 0)[1]) %>%
  ungroup() %>%
  left_join(site_year, by= c("SITE", "Year")) %>%
  group_by(sample.yr, SITE, SPECIES, Plot.Area.M2, sampled_plots_period, age_class_yr) %>%
  summarize(num_dead = n(), num_dead_pp = num_dead/mean(sampled_plots_period), num_dead_m2 = num_dead_pp/mean(Plot.Area.M2)) %>%
  ungroup() 


# Graph the mortality
ggplot(site_mortality, aes(sample.yr, num_dead_m2, color = age_class_yr))+
  geom_jitter(alpha=0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

# Join these dead with the density estimate with all of the trees at a site
site.mortality2 <- plot_data.2 %>%
  left_join(site_mortality) %>%
  mutate(num_dead_m2 = replace_na(num_dead_m2, 0), 
         prop_dead_m2 = num_dead_m2/no_trees_m2)

ggplot(site.mortality2, aes(sample.yr, prop_dead_m2, color = age_class_yr))+
  geom_jitter(alpha=0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)


# Large mortality
large_trees <- data.25 %>%
  filter(DBH>=10) 

large_mortality <- data.25 %>%
  filter(UniqueID%in% large_trees$UniqueID) %>%
  ungroup() %>%
  arrange(UniqueID, Year) %>%
  group_by(UniqueID) %>%
  # We look for the first index (row number) where DA is 0
  # but only if that index is NOT 1.
  filter(which(DA == 0)[1] > 1) %>%
  # Now that we've narrowed it to the right trees, 
  # we isolate that specific first-death row.
  slice(which(DA == 0)[1]) %>%
  ungroup() %>%
  left_join(site_year, by= c("SITE", "Year")) %>%
  group_by(sample.yr, SITE, SPECIES, Plot.Area.M2, sampled_plots_period, age_class_yr) %>%
  summarize(num_dead = n(), num_dead_pp = num_dead/mean(sampled_plots_period), num_dead_m2 = num_dead_pp/mean(Plot.Area.M2)) %>%
  ungroup() 


# Graph the mortality
ggplot(large_mortality, aes(sample.yr, num_dead_m2, color = age_class_yr))+
  geom_jitter(alpha=0.4)+
  theme_bw()+
  ggtitle("Large Tree Mortality")+
  facet_wrap(~SPECIES)


#msall mortality
small_trees <- data.25 %>%
  group_by(UniqueID) %>%
  summarize(max = max(DBH, na.rm=TRUE))%>%
  filter(max<10) 

small_mortality <- data.25 %>%
  filter(UniqueID %in% small_trees$UniqueID) %>%
  ungroup() %>%
  arrange(UniqueID, Year) %>%
  group_by(UniqueID) %>%
  # We look for the first index (row number) where DA is 0
  # but only if that index is NOT 1.
  filter(which(DA == 0)[1] > 1) %>%
  # Now that we've narrowed it to the right trees, 
  # we isolate that specific first-death row.
  slice(which(DA == 0)[1]) %>%
  ungroup() %>%
  left_join(site_year, by= c("SITE", "Year")) %>%
  group_by(sample.yr, SITE, SPECIES, Plot.Area.M2, sampled_plots_period, age_class_yr) %>%
  summarize(num_dead = n(), num_dead_pp = num_dead/mean(sampled_plots_period), num_dead_m2 = num_dead_pp/mean(Plot.Area.M2)) %>%
  ungroup() 

# Graph the mortality
ggplot(small_mortality, aes(sample.yr, num_dead_m2, color = age_class_yr))+
  geom_jitter(alpha=0.4)+
  theme_bw()+
  ggtitle("Small Tree Mortality")+
  facet_wrap(~SPECIES)


```


# Plot level BA deets
- note that BA includes all TREES not limited to dead/alive - should maybe update to just be alive....
```{r}
ba.plot <- data.25%>%
  filter(DA == 1) %>%
  mutate(BA = 0.00007854 * (DBH)^2) %>%
  ungroup() %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot.Area.M2) %>%
  summarize(n=n(), num_alive = sum(DA), plot_BA = sum(BA)) %>%
  left_join(site_year, by = c("SITE", "Year"))

# site level summary
ba.site <- ba.plot %>%
  ungroup()%>%
  group_by(sample.yr, SITE, SPECIES, sampled_plots_period, Plot.Area.M2, age_class_yr) %>%
  summarize(no_trees = sum(n), no_alive = sum(num_alive), sample = mean(sampled_plots), plot.area = mean(Plot.Area.M2), no_trees_pp = no_trees/sample, ba = sum(plot_BA, na.rm = TRUE), ba_pp = ba/sample, ba_m2 = ba_pp/plot.area)

ba.site.wide<- ba.site %>%
  select(-c(no_trees, no_alive, sample, ba, no_trees_pp, ba_pp))%>%
  ungroup()%>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("ba_m2"), values_fill=0) %>%
  mutate(sum_ba = POPBAL + POPTRE + BETNEO +PICGLA + PICEA + PICMAR + LARLAR, decid_ba = POPBAL + POPTRE + BETNEO, con_ba = PICGLA + PICEA + PICMAR + LARLAR) 

ggplot(ba.site.wide, aes(sample.yr, sum_ba, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  facet_wrap(~age_class_yr)

ggplot(ba.site.wide, aes(sample.yr, decid_ba, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  facet_wrap(~age_class_yr)

ggplot(ba.site.wide, aes(sample.yr, con_ba, color=age_class_yr))+
  geom_point()+
  theme_bw() + 
  facet_wrap(~age_class_yr)

ggplot(data = subset(ba.site, ba.site$SPECIES %!in% c("LARLAR", "PICEA")), aes(sample.yr, ba_m2, color = age_class_yr))+
  geom_jitter(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)
  
```
# BA and Density

```{r}
# taking site level BA and density for each sampling year
ba.dens <- left_join(ba.site.wide, plot_wide.2)
#str(ba.dens)

ggplot(ba.dens, aes(sum_ba, tot_alive))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")


ggplot(ba.dens, aes(sum_ba, decid_alive))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")

ggplot(ba.dens, aes(sum_ba, con_alive))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age_class_yr, scales="free")

```


# BA and Growth Rate

```{r}
ba.dens.gr <- left_join(ba.site, site_growth)
ba.dens.gr.2 <- ba.dens.gr %>%
  filter(!is.na(mean_AG))

#str(ba.dens.gr.2)

ggplot(ba.dens.gr.2, aes(ba_m2, max_dbh, color = age_class_yr)) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

ggplot(ba.dens.gr.2, aes(ba_m2, max_dbh, color = (no_trees_pp/plot.area ))) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

ggplot(ba.dens.gr.2, aes(ba_m2, mean_AG, color = age_class_yr)) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

ggplot(ba.dens.gr.2, aes(ba_m2, mean_AG, color = (no_trees_pp/plot.area ))) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

ggplot(ba.dens.gr.2, aes(ba_m2, mean_RBAG, color = age_class_yr)) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)

ggplot(ba.dens.gr.2, aes(ba_m2, mean_RGR, color = age_class_yr)) +
  geom_point(alpha = 0.4)+
  theme_bw()+
  facet_wrap(~SPECIES)
```



