---
title: "RSN_Data_exploration"
author: "OLH"
date: "2026-02-09"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```

# Summary of Q's
- many of the sites in the data don't have associated data in the site_info; these mostly look like older sites (i.e., first measurement is in 89 or 90s) and seem to be LTER sites like floodplain and upland, so starting with FP or UP
- trailing spaces in the SITE column in data; removing the trailing space

# Downlaod data
```{r}

# create data folders
dir.create(file.path("2026 Data"), showWarnings = F)
dir.create(file.path("2026 Data", 'RSN_Site'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data', 'Updated'), showWarnings = F)

# Identify desired files to download
site_info_file <- "RSN_Master_Site.csv"
qaqc_data <- "RSN_Inventory_QAQC_Final2.csv"

# Download data files
# site info
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1PWk45RgTK47B-PX71QG6BazpJyAtidQ1")) %>% 
  dplyr::filter(name == site_info_file) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "RSN_Site", .$name))

# qaqc data
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1VR7WRVcF3behOWDoXtuFlN9wK0M24jOb")) %>% 
  dplyr::filter(name == qaqc_data) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "QAQC_Data", "Updated", .$name))

# Read in site info and data files
site_info <- read.csv(file = file.path("2026 Data", 'RSN_Site', 'site_metadata_OLH.csv'))

data <-  read.csv(file = file.path("2026 Data", 'QAQC_Data','Updated', qaqc_data))

```

# Length and No of measurements
```{r}
unique(data$SITE)
data_summary <- data %>%
  # make a column for whether or not tree is above or below 2.5 cm
  mutate(over2.5 = ifelse(DBH>2.5, 1, 0)) %>%
  # make a column that indicates, dead, live, or seedling less than 2.5 cm
  mutate(tree_status = case_when(
    over2.5 == 1 & DA == 1 ~ as.character("Live"), 
    over2.5 == 0 & DA ==1 ~ as.character("Seedling"),
    DA == 0 ~ as.character("Dead"),
    DBH_raw == -8888 ~ as.character("Missing"), 
    DBH_raw %in% c(-9999, -7777) ~ as.character("Dead"),
    TRUE ~ "Missing"
  )) %>%
  group_by(SITE, Year, tree_status) %>%
  summarize(no_plots = n_distinct(PLOT), no_trees = n_distinct(TREE)) %>%
  pivot_wider(names_from = tree_status, values_from = c(no_plots, no_trees), values_fill = 0) %>%
  ungroup()%>%
  group_by(SITE) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(live_diff = no_trees_Live - lag(no_trees_Live))

# write.csv(data_summary,'/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_year_tree_summary.csv', row.names = FALSE)


# 1. Open the PDF file
# pdf("/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/Site_Trends.pdf", width = 10, height = 7)

# 2. Get a list of all sites
plot_ready <- data_summary %>%
  # Select only the columns needed for the graph
  select(SITE, Year, no_trees_Live, no_trees_Dead, no_trees_Seedling) %>%
  # Pivot so 'Live' and 'Dead' are in one column
  pivot_longer(cols = c(no_trees_Live, no_trees_Dead, no_trees_Seedling), 
               names_to = "Status", 
               values_to = "Count") %>%
  # Clean up labels: "no_trees_Live" becomes "Live"
  mutate(Status = str_remove(Status, "no_trees_"))
site_list <- unique(plot_ready$SITE)

for (s in site_list) {
  p <- plot_ready %>%
    filter(SITE == s) %>%
    ggplot(aes(x = Year, y = Count, color = Status, group = Status)) +
    geom_line(size = 1.2) +
    geom_point(size = 3) +
    scale_color_manual(values = c("Live" = "#2ecc71", "Dead" = "#e74c3c", "Seedling" = "skyblue")) +
    labs(title = paste("Tree Density Trends - Site:", s),
         subtitle = "Comparing Live vs. Dead stems over time",
         x = "Year", y = "Number of Trees") +
    theme_minimal()
  
  print(p) # This "prints" the plot as a new page in the PDF
}

```

# Create a table for the years and sites

```{r}
str(data_summary)
site_yrs <- data %>%
  select(SITE, Year, PLOT) %>%
  distinct() %>%
  group_by(SITE, Year) %>%
  summarize(n_plot = n_distinct(PLOT)) %>%
  arrange(Year) %>%
  mutate(present = 1) %>%
  left_join(site_info, by = "SITE")

# # divide sites
# older <- site_yrs %>%
#   filter(Year < 2013)
unique(site_yrs$age_class)
mature <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Mature"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Mature Stands")
mature

inter <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Intermediate"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Intermediate Stands")
inter

young <- ggplot(data = subset(site_yrs, site_yrs$age_class == "Young"), aes(SITE, Year)) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Young Stands")
young

# unk <- ggplot(data = subset(site_yrs, is.na(site_yrs$age_class)), aes(SITE, Year)) + 
#   geom_point() +
#   theme_bw()+
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
#   ggtitle("Unknown Stands")
# unk

unique(data$Year)
```

# No plots (v1)
- this was done on older data...

```{r}
# unique(data$SITE)
# data_yrs <- data %>%
#   # remove any trailing spaces
#   mutate(SITE = str_trim(SITE, side = "right"))%>%
#   select(SITE, PLOT, Year) %>%
#   distinct() 
# 
# ggplot(data_yrs, aes(SITE, Year))+
#   geom_point()+
#   theme_minimal()+
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
# 
# # how many plots in each year
# no_plots <- data %>%
#   # remove any trailing spaces
#   mutate(SITE = str_trim(SITE, side = "right"))%>%
#   select(SITE, PLOT, Year) %>%
#   distinct() %>%
#   group_by(SITE, Year) %>%
#   summarize(no_plots = n_distinct(PLOT)) %>%
#   rename(sitename = SITE) 
# 
# no_plots.filter <- no_plots %>%
#   group_by(sitename) %>%
#   filter(all(no_plots %in% c(12, 30))) %>%
#   ungroup() %>%
#   select(sitename, no_plots) %>%
#   distinct()
# 
# no_plots.filter.opp <- no_plots %>%
#   filter(sitename %!in% no_plots.filter$sitename) %>% 
#   select(sitename) %>%
#   distinct()
  
# write.csv(no_plots, '/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/no_plots.csv')
```
## Histograms of dbh
```{r}
data.dbh <- data %>%
  filter(!is.na(DBH))
ggplot(data.dbh, aes(DBH)) +
  geom_histogram()+
  theme_bw()+
  facet_wrap(~SPECIES, scales="free")
  
salbeb <- data %>%
  filter(SPECIES == "SALBEB")
  

```

# Plots missing site data
- have an updated metadata file that now will have all of the sites

```{r}
# site_missing <- data %>%
#   rename(sitename = "SITE") %>%
#   left_join(site_info, by = "sitename") %>%
#   filter(is.na(age)) %>%
#   select(sitename) %>%
#   distinct()
# 
# # write.csv(site_missing,'/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_missing.csv', row.names = FALSE)
```



## Data cleaning round 2 - after the updates
- most of initial concerns have been addressed
```{r}
unique(data$SPECIES)
data.2 <- data %>%
  # trim the site to be the right
   # mutate(SITE = str_trim(SITE, side = "right"))%>%
  # get rid of the only salix species
   filter(SPECIES != "SALBEB") 
  # update unknowns that seem to be a capitalization issue
  # mutate(SPECIES = case_when(
  #          SPECIES_raw == "popbal" ~ 'POPBAL', 
  #          SPECIES_raw == "Popbal"~ 'POPBAL', 
  #          SPECIES_raw == "betneo"~ 'BETNEO',
  #          SPECIES_raw == "Betneo"~ 'BETNEO',
  #          SPECIES_raw == "picgla"~ 'PICGLA',
  #          SPECIES_raw == "Popbal      "~ 'POPBAL',
  #          SPECIES_raw == "picgla "~ 'PICGLA',
  #          TRUE ~ SPECIES
  #        ))

```

# DBH THROUGH TIME
- tree level
```{r}
str(data.2)
# try getting rid of the NAs
data.dbh <- data.2 %>%
  filter(!is.na(DBH)) %>%
  # can not do this too, but get rid of really small trees for consistency
  filter(DBH> 2.4) %>%
  group_by(UniqueID) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    DBH_rel2 = pi * (DBH/2)^2,
    BA = 0.00007854 * (DBH)^2,
    DBH_diff = DBH - lag(DBH),
    DBH_rel_diff = ((DBH_rel2 - lag(DBH_rel2))/lag(DBH_rel2))*100,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = DBH_diff / Year_diff,
    DBH_relativized_BA = DBH_rel_diff/Year_diff,
    BA_growth = (BA_diff/Year_diff)  * 10000,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    Year_before = lag(Year)
  ) %>%
  #filter(!is.na(Annual_Growth)) %>%
  left_join(site_info, by = "SITE")

```

## Some initial viz

### PicMAR
```{r, fig.width=12, fig.height=6, fig.retina=2}
picmar <- data.dbh %>%
  filter(SPECIES == "PICMAR")
picmar.rgr <- ggplot(data = picmar, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.rgr

picmar.rba <- ggplot(data = subset(picmar, picmar$DBH_relativized_BA < 100), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.rba

picmar.ba <- ggplot(data = picmar, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.ba

picmar.ann <- ggplot(data = picmar, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picmar.ann
```

### PICGLA

```{r, fig.width=12, fig.height=6, fig.retina=2}

picgla <- data.dbh %>%
  filter(SPECIES == "PICGLA")
picgla.rgr <- ggplot(data = picgla, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.rgr

picgla.rba <- ggplot(data = subset(picgla, picgla$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.rba

picgla.ba <- ggplot(data = picgla, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.ba

picgla.ann <- ggplot(data = picgla, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
picgla.ann

```

### POPTRE

```{r, fig.width=12, fig.height=6, fig.retina=2}

poptre <- data.dbh %>%
  filter(SPECIES == "POPTRE")
poptre.rgr <- ggplot(data = poptre, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.rgr

poptre.rba <- ggplot(data = poptre, aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.rba

poptre.ba <- ggplot(data = poptre, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.ba

poptre.ann <- ggplot(data = poptre, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
poptre.ann

```
### BETNEO
```{r, fig.width=12, fig.height=6, fig.retina=2}

betneo <- data.dbh %>%
  filter(SPECIES == "BETNEO")
betneo.rgr <- ggplot(data = betneo, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.rgr

betneo.rba <- ggplot(data = subset(betneo, betneo$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.rba

betneo.ba <- ggplot(data = betneo, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.ba

betneo.ann <- ggplot(data = betneo, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
betneo.ann

```

### POPBAL

```{r, fig.width=12, fig.height=6, fig.retina=2}

popbal <- data.dbh %>%
  filter(SPECIES == "POPBAL")
popbal.rgr <- ggplot(data = popbal, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.rgr

popbal.rba <- ggplot(data = subset(popbal, popbal$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.rba

popbal.ba <- ggplot(data = popbal, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.ba

popbal.ann <- ggplot(data = popbal, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
popbal.ann

```


##LARLAR

```{r, fig.width=12, fig.height=6, fig.retina=2}

larlar <- data.dbh %>%
  filter(SPECIES == "LARLAR")
larlar.rgr <- ggplot(data = larlar, aes(Year, RGR, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.rgr

larlar.rba <- ggplot(data = subset(larlar, larlar$DBH_relativized_BA < 1000), aes(Year, DBH_relativized_BA, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.rba

larlar.ba <- ggplot(data = larlar, aes(Year, BA_growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.ba

larlar.ann <- ggplot(data = larlar, aes(Year, Annual_Growth, color = Year_before))+
  geom_jitter()+
  theme_bw()+
  facet_wrap(~age_class, scales="free_x")
larlar.ann

```

# Tree growth summary table through time
- going to summarize on the decade, by tree species and age class
- want a count of the number of unique trees too

```{r}
str(data.dbh)
# function to floor decade 
floor_decade    = function(value){ return(value - value %% 10) }

individ.summary <- data.dbh %>%
  ungroup() %>%
  # make a decade column - floor
  mutate(decade = floor_decade(Year)) %>%
  filter(!is.na(DBH_diff)) %>%
  group_by(decade, SPECIES, age_class) %>%
  summarize(n_tree = n_distinct(UniqueID), n_site = n_distinct(SITE), 
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()), 
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_BG = mean(BA_growth), median_BG = median(BA_growth), se_BG = sd(BA_growth)/sqrt(n()),
            mean_RBAG = mean(DBH_relativized_BA), median_RBAG = median(DBH_relativized_BA), se_RBAG = sd(DBH_relativized_BA)/sqrt(n())
            ) %>%
  arrange(SPECIES, decade, age_class) 
  
## same code but by year
individ.summary.yr <- data.dbh %>%
  ungroup() %>%
  filter(!is.na(DBH_diff)) %>%
  group_by(Year, SPECIES, age_class) %>%
  summarize(n_tree = n_distinct(UniqueID), n_site = n_distinct(SITE), 
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()), 
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_BG = mean(BA_growth), median_BG = median(BA_growth), se_BG = sd(BA_growth)/sqrt(n()),
            mean_RBAG = mean(DBH_relativized_BA), median_RBAG = median(DBH_relativized_BA), se_RBAG = sd(DBH_relativized_BA)/sqrt(n())
            ) %>%
  arrange(SPECIES, Year, age_class) 
  


# Filter for visuals = get rid of picea, larlar
individ.sum.vis <- individ.summary %>%
  # tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR")) %>%
  # get rid of obs with only 1 site
  filter(n_site >1) %>%
  # requrie at least 10 trees?
  filter(n_tree >10)
  

individ.sum.vis.yr <- individ.summary.yr %>%
  # tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR")) %>%
  # get rid of obs with only 1 site
  #filter(n_site >1) %>%
  # requrie at least 10 trees?
  filter(n_tree >10)
```

# Visuals of individ growth over time and by age class - DECADE
```{r}
ggplot(individ.sum.vis, aes(decade, n_tree, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis, aes(decade, n_site, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis, aes(decade, mean_dbh, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh - se_dbh, ymax = mean_dbh + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis, aes(decade, mean_AG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG - se_AG, ymax = mean_AG + se_AG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_AG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_AG - se_AG, ymax = median_AG + se_AG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(individ.sum.vis, aes(decade, mean_RGR, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR - se_RGR, ymax = mean_RGR + se_RGR),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

# 
# ggplot(individ.sum.vis, aes(decade, median_RGR, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_RGR - se_RGR, ymax = median_RGR + se_RGR),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis, aes(decade, mean_BG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG - se_BG, ymax = mean_BG + se_BG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_BG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_BG - se_BG, ymax = median_BG + se_BG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(individ.sum.vis, aes(decade, mean_RBAG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG - se_RBAG, ymax = mean_RBAG + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_RBAG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_RBAG - se_RBAG, ymax = median_RBAG + se_RBAG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")

```


# Visual by YEAR
```{r}
ggplot(individ.sum.vis.yr, aes(Year, n_tree, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis.yr, aes(Year, n_site, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis.yr, aes(Year, mean_dbh, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_dbh - se_dbh, ymax = mean_dbh + se_dbh),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis.yr, aes(Year, mean_AG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_AG - se_AG, ymax = mean_AG + se_AG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_AG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_AG - se_AG, ymax = median_AG + se_AG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(individ.sum.vis.yr, aes(Year, mean_RGR, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RGR - se_RGR, ymax = mean_RGR + se_RGR),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")

# 
# ggplot(individ.sum.vis, aes(decade, median_RGR, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_RGR - se_RGR, ymax = median_RGR + se_RGR),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")

ggplot(individ.sum.vis.yr, aes(Year, mean_BG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_BG - se_BG, ymax = mean_BG + se_BG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_BG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_BG - se_BG, ymax = median_BG + se_BG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")


ggplot(individ.sum.vis.yr, aes(Year, mean_RBAG, group = age_class, color=age_class))+
  geom_point(position = position_dodge(width = 0.6), size = 1) +
  geom_errorbar(
    aes(ymin = mean_RBAG - se_RBAG, ymax = mean_RBAG + se_RBAG),
    width = 0.2,
    position = position_dodge(width = 0.6)
  )+
  theme_bw()+
  facet_wrap(~SPECIES, nrow=3, scales="free")


# ggplot(individ.sum.vis, aes(decade, median_RBAG, group = age_class, color=age_class))+
#   geom_point(position = position_dodge(width = 0.6), size = 1) +
#   geom_errorbar(
#     aes(ymin = median_RBAG - se_RBAG, ymax = median_RBAG + se_RBAG),
#     width = 0.2,
#     position = position_dodge(width = 0.6)
#   )+
#   theme_bw()+
#   facet_wrap(~SPECIES, nrow=3, scales="free")
```

# Understand missing trees
- There are 859 trees that end up going missing that were measured and their last measurement does not indicate that they are dead
- 47 site-yeras with more than 1 tree missing, and only 20 with more than 5. going to look into this more
- want to make sure that any tree that is going to be recorded as dead had a record before...?
- there are 3116 observations of only NA, but its only 1972 distinct trees - I think a lot of these are because they had DBH less than 2.5 in iniital obs, because if I remove that requirement, then there are 154 obs corresponding to 85 trees - this doesn't really affect the last measurement, but will be important to remove (?) for demographic surveys

```{r}
# get rid of the small trees
data.25 <- data %>%
  filter(is.na(DBH) | DBH > 2.4)

# see which trees only have an NA
na.trees <- data.25 %>%
  group_by(UniqueID) %>%
  filter(all(is.na(DBH))) #%>%
  #distinct(UniqueID)

# remove the na trees only
data.25 <- data.25 %>%
  filter(UniqueID %!in% na.trees$UniqueID)

# Derive site visits from tree data if no separate visit table
site_visits <- data.25 %>%
  distinct(SITE, Year)

# --- Step 1: Find the last occasion each tree was recorded ---
last_seen <- data.25 %>%
  group_by(UniqueID, SITE) %>%
  summarise(last_occasion = max(Year), .groups = "drop")

# --- Step 2: Get all site visits that occurred AFTER each tree's last record ---
# Cross join trees to visits, then filter to later visits
later_visits <- last_seen %>%
  left_join(site_visits, by = "SITE") %>%       # all visits for that site
  filter(Year > last_occasion) %>%                  # only visits after last record
  # mutate to create a new column that has the difference between teh year and last occassion adn then for each site/tree, select the smallest value
  mutate(diff = Year - last_occasion) %>%
  group_by(UniqueID, SITE) %>%
  slice_min(diff, n=1)
  
# select the trees identified in the data file, and filter that those observations are not dead
data.select <- data.25 %>%
  filter(UniqueID %in% later_visits$UniqueID) %>%
  group_by(UniqueID, SITE) %>%
  slice_max(Year, n =1) %>%
  filter(DA == 1)

site_miss_tree <- data.select %>%
  ungroup() %>%
  group_by(SITE, Year) %>%
  dplyr::count() %>%
  filter(n>5)
```


# Plot and Site level details...?

## Plot basal area over time

## Density 

```{r}
plot_data <- data %>%
  # trim the site to be the right
   mutate(SITE = str_trim(SITE, side = "right"))%>%
  # get rid of the only salix species
   filter(SPECIES != "SALBEB") %>%
  # if it has no trees or has fallen - make -999 0
  mutate(DA = ifelse(DA == -9999, 0, DA)) %>%
  mutate(DA = ifelse(DA == -999, 0, DA)) %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot_M2) %>%
  summarize(n=n(), num_alive = sum(DA)) %>%
  mutate(prop_alive = (num_alive/n)*100) 

ggplot(plot_data, aes(Year, prop_alive))+
  geom_point()+
  theme_bw() +
  facet_wrap(~SPECIES)

# make it wider to look at full plot level dynamics
plot_wide <- plot_data %>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("n", "num_alive", "prop_alive"), values_fill=0) %>%
  mutate(tot_trees = n_POPBAL + n_POPTRE + n_PICGLA + n_PICEA + n_PICMAR + n_UNK + n_LARLAR + n_BETNEO, 
         tot_alive = num_alive_POPBAL + num_alive_POPTRE + num_alive_PICGLA + num_alive_PICEA + num_alive_PICMAR + num_alive_UNK + num_alive_LARLAR + num_alive_BETNEO, 
         prop_alive = tot_alive/tot_trees) %>%
  rename(sitename= "SITE")%>%
  left_join(site_info, by = c("sitename")) %>%
  mutate(age = ifelse(is.na(age), "mature", age))

ggplot(plot_wide, aes(Year.x, prop_alive))+
  geom_point()+
  theme_bw() + 
  facet_wrap(~age)

# Summarize the site and counts
site_plots <- plot_wide %>%
  ungroup()%>%
  group_by(sitename, Year.x) %>%
  summarize(no_plots = length(unique(PLOT))) %>%
  select(-Year.x)%>%
  distinct() %>%
  mutate(binary = 1) %>%
  group_by(sitename) %>%
  summarize(sum= sum(binary)) %>%
  filter(sum >1)

site_plots2 <- plot_wide %>%
  ungroup()%>%
  group_by(sitename, Year.x) %>%
  summarize(no_plots = length(unique(PLOT))) %>%
  left_join(site_plots)

ggplot(site_plots2, aes(sitename, no_plots))+
  geom_point()+
  theme_bw()
  
```

## Plot level BA deets
