---
title: "RSN_Data_exploration"
author: "OLH"
date: "2026-02-09"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot,
)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```

# Summary of Q's
- many of the sites in the data don't have associated data in the site_info; these mostly look like older sites (i.e., first measurement is in 89 or 90s) and seem to be LTER sites like floodplain and upland, so starting with FP or UP
- trailing spaces in the SITE column in data; removing the trailing space

# Downlaod data
```{r}

# create data folders
dir.create(file.path("2026 Data"), showWarnings = F)
dir.create(file.path("2026 Data", 'RSN_Site'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data'), showWarnings = F)

# Identify desired files to download
site_info_file <- "RSN_Master_Site.csv"
qaqc_data <- "RSN_Inventory_QAQC_Final.csv"

# Download data files
# site info
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1PWk45RgTK47B-PX71QG6BazpJyAtidQ1")) %>% 
  dplyr::filter(name == site_info_file) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "RSN_Site", .$name))

# qaqc data
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1mT036D-xu_XPbJGupVoAKH2DK982oGxx")) %>% 
  dplyr::filter(name == qaqc_data) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "QAQC_Data", .$name))

# Read in site info and data files
site_info <- read.csv(file = file.path("2026 Data", 'RSN_Site', site_info_file))

data <-  read.csv(file = file.path("2026 Data", 'QAQC_Data', qaqc_data))

```

# Length and No of measurements
```{r}
data_summary <- data %>%
  # remove any trailing spaces
  mutate(SITE = str_trim(SITE, side = "right"))%>%
  select(SITE, PLOT, Year) %>%
  distinct() %>%
  group_by(SITE) %>%
  summarize(no_yrs = n_distinct(Year), no_plots = n_distinct(PLOT), first_year = min(Year), last_year = max(Year)) %>%
  rename(sitename = SITE) %>%
  left_join(site_info)

missing <- data %>%
  filter(SITE %!in% data_summary$sitename)
unique(data$SITE)
data_yrs <- data %>%
  # remove any trailing spaces
  mutate(SITE = str_trim(SITE, side = "right"))%>%
  select(SITE, PLOT, Year) %>%
  distinct() 

ggplot(data_yrs, aes(SITE, Year))+
  geom_point()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```
## Histograms of dbh
```{r}
data.dbh <- data %>%
  filter(!is.na(DBH))
ggplot(data.dbh, aes(DBH)) +
  geom_histogram()+
  theme_bw()+
  facet_wrap(~SPECIES, scales="free")
  
salbeb <- data %>%
  filter(SPECIES == "SALBEB")
  

```

## Need to do some cleaning
```{r}
data.2 <- data %>%
  # trim the site to be the right
   mutate(SITE = str_trim(SITE, side = "right"))%>%
  # get rid of the only salix species
   filter(SPECIES != "SALBEB") %>%
  # update unknowns that seem to be a capitalization issue
  mutate(SPECIES = case_when(
           SPECIES_raw == "popbal" ~ 'POPBAL', 
           SPECIES_raw == "Popbal"~ 'POPBAL', 
           SPECIES_raw == "betneo"~ 'BETNEO',
           SPECIES_raw == "Betneo"~ 'BETNEO',
           SPECIES_raw == "picgla"~ 'PICGLA',
           SPECIES_raw == "Popbal      "~ 'POPBAL',
           SPECIES_raw == "picgla "~ 'PICGLA',
           TRUE ~ SPECIES
         ))

# check which unknowns are left
unk <- data.2 %>%
  filter(SPECIES == "UNK")
unique(unk$SPECIES_raw)
```

# DBH THROUGH TIME
- tree level
```{r}
str(data.2)
# try getting rid of the NAs
data.dbh <- data.2 %>%
  filter(!is.na(DBH)) %>%
  group_by(UniqueID) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    DBH_rel2 = pi * (DBH/2)^2,
    BA = 0.00007854 * (DBH)^2,
    DBH_diff = DBH - lag(DBH),
    DBH_rel_diff = ((DBH_rel2 - lag(DBH_rel2))/lag(DBH_rel2))*100,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = DBH_diff / Year_diff,
    DBH_relativized_BA = DBH_rel_diff/Year_diff,
    BA_growth = (BA_diff/Year_diff)  * 10000,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    Year_before = lag(Year)
  ) %>%
  #filter(!is.na(Annual_Growth)) %>%
  rename(sitename = "SITE") %>%
  left_join(site_info, by = "sitename") %>%
  mutate(age = ifelse(is.na(age), "mature", age))


```

## Some initial viz

### PicMAR
```{r, fig.width=12, fig.height=6, fig.retina=2}
picmar <- data.dbh %>%
  filter(SPECIES == "PICMAR")
picmar.rgr <- ggplot(data = picmar, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picmar.rgr

picmar.rba <- ggplot(data = subset(picmar, picmar$DBH_relativized_BA < 100), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picmar.rba

picmar.ba <- ggplot(data = picmar, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picmar.ba

picmar.ann <- ggplot(data = picmar, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picmar.ann
```

### PICGLA

```{r, fig.width=12, fig.height=6, fig.retina=2}

picgla <- data.dbh %>%
  filter(SPECIES == "PICGLA")
picgla.rgr <- ggplot(data = picgla, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picgla.rgr

picgla.rba <- ggplot(data = subset(picgla, picgla$DBH_relativized_BA < 1000), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picgla.rba

picgla.ba <- ggplot(data = picgla, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picgla.ba

picgla.ann <- ggplot(data = picgla, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
picgla.ann

```

### POPTRE

```{r, fig.width=12, fig.height=6, fig.retina=2}

poptre <- data.dbh %>%
  filter(SPECIES == "POPTRE")
poptre.rgr <- ggplot(data = poptre, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
poptre.rgr

poptre.rba <- ggplot(data = subset(poptre, poptre$DBH_relativized_BA < 6000), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
poptre.rba

poptre.ba <- ggplot(data = poptre, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
poptre.ba

poptre.ann <- ggplot(data = poptre, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
poptre.ann

```
### BETNEO
```{r, fig.width=12, fig.height=6, fig.retina=2}

betneo <- data.dbh %>%
  filter(SPECIES == "BETNEO")
betneo.rgr <- ggplot(data = betneo, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
betneo.rgr

betneo.rba <- ggplot(data = subset(betneo, betneo$DBH_relativized_BA < 1000), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
betneo.rba

betneo.ba <- ggplot(data = betneo, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
betneo.ba

betneo.ann <- ggplot(data = betneo, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
betneo.ann

```

### POPBAL

```{r, fig.width=12, fig.height=6, fig.retina=2}

popbal <- data.dbh %>%
  filter(SPECIES == "POPBAL")
popbal.rgr <- ggplot(data = popbal, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
popbal.rgr

popbal.rba <- ggplot(data = subset(popbal, popbal$DBH_relativized_BA < 1000), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
popbal.rba

popbal.ba <- ggplot(data = popbal, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
popbal.ba

popbal.ann <- ggplot(data = popbal, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
popbal.ann

```


##LARLAR

```{r, fig.width=12, fig.height=6, fig.retina=2}

larlar <- data.dbh %>%
  filter(SPECIES == "LARLAR")
larlar.rgr <- ggplot(data = larlar, aes(Year.x, RGR, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
larlar.rgr

larlar.rba <- ggplot(data = subset(larlar, larlar$DBH_relativized_BA < 1000), aes(Year.x, DBH_relativized_BA, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
larlar.rba

larlar.ba <- ggplot(data = larlar, aes(Year.x, BA_growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
larlar.ba

larlar.ann <- ggplot(data = larlar, aes(Year.x, Annual_Growth, color = Year_before))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age, scales="free")
larlar.ann

```


# Plot and Site level details...?

## Density 

```{r}
plot_data <- data %>%
  # trim the site to be the right
   mutate(SITE = str_trim(SITE, side = "right"))%>%
  # get rid of the only salix species
   filter(SPECIES != "SALBEB") %>%
  # if it has no trees or has fallen - make -999 0
  mutate(DA = ifelse(DA == -9999, 0, DA)) %>%
  mutate(DA = ifelse(DA == -999, 0, DA)) %>%
  group_by(Year,SITE,PLOT, SPECIES, Plot_M2) %>%
  summarize(n=n(), num_alive = sum(DA)) %>%
  mutate(prop_alive = (num_alive/n)*100) 

ggplot(plot_data, aes(Year, prop_alive))+
  geom_point()+
  theme_bw() +
  facet_wrap(~SPECIES)

# make it wider to look at full plot level dynamics
plot_wide <- plot_data %>%
  pivot_wider(names_from = SPECIES, 
              values_from=c("n", "num_alive", "prop_alive"), values_fill=0) %>%
  mutate(tot_trees = n_POPBAL + n_POPTRE + n_PICGLA + n_PICEA + n_PICMAR + n_UNK + n_LARLAR + n_BETNEO, 
         tot_alive = num_alive_POPBAL + num_alive_POPTRE + num_alive_PICGLA + num_alive_PICEA + num_alive_PICMAR + num_alive_UNK + num_alive_LARLAR + num_alive_BETNEO, 
         prop_alive = tot_alive/tot_trees) %>%
  rename(sitename= "SITE")%>%
  left_join(site_info, by = c("sitename")) %>%
  mutate(age = ifelse(is.na(age), "mature", age))

ggplot(plot_wide, aes(Year.x, prop_alive))+
  geom_point()+
  theme_bw() + 
  facet_wrap(~age)

# Summarize the site and counts
site_plots <- plot_wide %>%
  ungroup()%>%
  group_by(SITE, Year) %>%
  summarize(no_plots = length(unique(PLOT))) %>%
  select(-Year)%>%
  distinct() %>%
  mutate(binary = 1) %>%
  group_by(SITE) %>%
  summarize(sum= sum(binary)) %>%
  filter(sum >1)

site_plots2 <- plot_wide %>%
  ungroup()%>%
  group_by(SITE, Year) %>%
  summarize(no_plots = length(unique(PLOT))) %>%
  left_join(site_plots)

ggplot(site_plots2, aes(SITE, no_plots))+
  geom_point()+
  theme_bw()
  
```

## Plot level BA deets
