---
title: "Merging_RSN_CAFI"
author: "OLH"
date: "2026-05-21"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```


# Read in Data
```{r}
## RSN Datasets
rsn <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/QAQC_Data/Updated/RSN_Inventory_QAQC_Final_2026-04-09.csv')

site.yrs <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_years.csv')

## CAFI Datasets
cafi <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/CAFI_Inventory_QAQC_Final.csv")
cafi.ages <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/cafi_loc_deciduous.csv")

## Metadata
pest <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/Pest_merge_check.csv')

```

# RSN Processing

## Data cleaning
- remove sites not in the RSN?
- Get rid of trees that are below 2.5 cm at DBH (Unless they shrink down)
- Get rid of trees that are only ever NA
- Remove one SALBEB obs

```{r}
# sites not in the RSN
non_rsn_sites <- c("FP0A", "FP0B","FP0C", "FP1A","FP1B","FP1C", "FP2A", "FP2B","FP2C",   "FP3A","FP3B","FP3C",'FP4A', 'FP4B', 'FP4C', 'FP5B', "UP1A", "UP1B","UP1C","UP2A", "UP2B", "UP2C", "UP3C", "UP3A", "UP3B")

# initial updates to RSN dataset
rsn.2 <- rsn %>%
  # remove salbeb
  filter(SPECIES != "SALBEB") %>%
  # remove sites not in the RSN
  filter(SITE %!in% non_rsn_sites)%>%
  # set up for removing small trees
  arrange(UniqueID, Year) %>% # Ensure data is in chronological order
  group_by(UniqueID) %>%
  # keep NA DBH, DBH > 2.5, or if the DBH exceeds 2.5, it will keep subsequent observatoins
  # this code above removes 4693 observations
  filter(is.na(DBH) | DBH >= 2.49 | cumany(DBH >= 2.49)) %>%
  ungroup()

# see which trees only have an NA - this is after the seedlings have been removed, so in many cases these had an intiial measurement that was NA and corresponded to somehting like 0.7cm
na.trees <- rsn.2%>%
  group_by(UniqueID) %>%
  filter(all(is.na(DBH))) #%>%
  #distinct(UniqueID)

# remove the na trees only
rsn.3 <- rsn.2 %>%
  filter(UniqueID %!in% na.trees$UniqueID) %>%
  # get rid of sampling rounds before the formal RSN initiation
  filter(Year> 2007)

```

## How many trees start off above 5
- There are a lot of trees over 5cm for first DBH - 624 trees i think and it is mostly FP5A and FP5D (430 is these two)
-I think for FP5A and FP5D, I think that we missed a sampling interval.... so these tries might have existed 
- taking care of that, i think its really 213 that are concerning in terms of over 5; really only over a handful that are over 6 - seem like about 200 trees in total and then only 73 if looking at 6cm

```{r}
early.trees <- rsn.3 %>%
  filter(Year < 2017)

late <- rsn.3 %>%
  filter(UniqueID %!in% early.trees$UniqueID) %>%
  filter(DBH > 5) #%>%
  # select(UniqueID) %>%
  # distinct()

ggplot(late, aes(DBH))+
  geom_histogram()+
  facet_wrap(~SPECIES, scales = "free")

# going to see if these late ones are measured twice?
late2 <- rsn.3 %>%
  filter(UniqueID %in% late$UniqueID) %>%
  filter(DBH < 6)

late3 <- rsn.3 %>%
  filter(UniqueID %in% late$UniqueID) %>%
  filter(UniqueID %!in% late2$UniqueID) %>%
  group_by(SITE) %>%
  summarize(n=n())

# Can run or not if wanting to get rid of these observations; right now its 73 trees that show up greater than 6 in the first obs round
# rsn.4 <- rsn.3 %>%
#   filter(UniqueID %!in% late3$UniqueID)
#   

```

## Clean up the damage and severity columsn - only done in the final sampling period
```{r}
unique(rsn.3$DAMAGE)
unique(rsn.3$LOCATION)
unique(rsn.3$SEVERITY)

# this code was done to identify all of the unique pests that were a thing
# unique.pests <- rsn.3 %>%
#     mutate(DAMAGE2 = case_when(
#     DAMAGE == "Girdled and Black rust" ~ "Girdled, Black rust", 
#     DAMAGE == "Leafroller Leafminer" ~ "Leafroller, Leafminer",
#     TRUE ~ DAMAGE
#   )) %>%
#   select(DAMAGE, DAMAGE2) %>%
#   distinct() %>%
#   mutate(DAMAGE_type = strsplit(DAMAGE, ",\\s*")) %>% 
#   unnest_wider(DAMAGE_type, names_sep = "_")

# prep teh pest dataset for RSN analysis
pests.2 <- pest %>%
  filter(Network == "RSN") 

rsn.4 <- rsn.3 %>%
  # there are a few categories that need to be fixed so that they separate 
  mutate(DAMAGE = case_when(
    DAMAGE == "Girdled and Black rust" ~ "Girdled, Black rust", 
    DAMAGE == "Leafroller Leafminer" ~ "Leafroller, Leafminer",
    TRUE ~ DAMAGE
  )) %>%
  mutate(DAMAGED = ifelse(DAMAGE == "", 0, 1),
    DAMAGE_type = strsplit(DAMAGE, ",\\s*")) %>% 
  unnest_wider(DAMAGE_type, names_sep = "_")%>%
  # Step 1 - needing to clean up the names based on the pests lookup table
  left_join(pests.2, by = c("DAMAGE_type_1" = "Common_Name")) %>%
  mutate(DAMAGE_1 = Damage) %>%
  select(-c(Network, Code, Damage)) %>%
  left_join(pests.2, by = c("DAMAGE_type_2" = "Common_Name")) %>%
  mutate(DAMAGE_2 = Damage) %>%
  select(-c(Network, Code, Damage)) %>%
  left_join(pests.2, by = c("DAMAGE_type_3" = "Common_Name")) %>%
  mutate(DAMAGE_3 = Damage) %>%
  select(-c(Network, Code, Damage, DAMAGE_type_1, DAMAGE_type_2, DAMAGE_type_3))%>%
  # Now will work on the severity
  mutate(SEVERITY = case_when(
    SEVERITY == "1.1" ~ "1,1",
    TRUE ~ SEVERITY
  )) %>%
  mutate(DAMAGE_severity = strsplit(SEVERITY, ",\\s*")) %>% 
  unnest_wider(DAMAGE_severity, names_sep = "_") %>%
  mutate(DAMAGE_severity_1 = case_when(
    DAMAGE_severity_1 == "4 (36-67%)" ~ "4",
    DAMAGE_severity_1 == "3 (16-35%)" ~ "3",
    DAMAGE_severity_1 == "1 (Trace to 5%)" ~ "1",
    DAMAGE_severity_1 == "5 (68-100%)" ~ "5",
    DAMAGE_severity_1 == "2 (6-15%)"  ~ "2",
    DAMAGE_severity_1 == "very severe" ~ "5",
    DAMAGE_severity_1 == "mo" ~ "3",
    DAMAGE_severity_1 == "mild" ~ "1",
    DAMAGE_severity_1 == "mi" ~ "1",
    DAMAGE_severity_1 == "12" ~ "2",
    TRUE ~ DAMAGE_severity_1
  ))%>%
  # Update damage to match the 1-4 scale of CAFI
  mutate(DAMSEV1= case_when(
    DAMAGE_severity_1 == "3" ~ "2", 
    DAMAGE_severity_1 == "4" ~ "3", 
    DAMAGE_severity_1 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_1
  ), 
  DAMSEV2 = case_when(
    DAMAGE_severity_2 == "3" ~ "2", 
    DAMAGE_severity_2 == "4" ~ "3", 
    DAMAGE_severity_2 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_2
  ), 
  DAMSEV3= case_when(
    DAMAGE_severity_3 == "3" ~ "2", 
    DAMAGE_severity_3 == "4" ~ "3", 
    DAMAGE_severity_3 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_3
  )
  ) %>%
  # Calculate max severity
  mutate(max_damage =pmax(DAMSEV1,DAMSEV2,DAMSEV3)) 
str(rsn.4)
```


## Merge this with the site years
```{r}
rsn.5 <- rsn.4 %>%
  left_join(site.yrs)
```

## Reduce to the necessary columns
- here is what i want: UniqueID, Year, SUBPLOT, Cycle (?), SITE, TREE, SPECIES, DBH, DA, TOP(?), Lean(?), Bowed(?), DOWN (?), Plot_area, Plot_size, no_plots, damage, damage_loc, damage_severity, QAQC_ACTION

```{r}
str(rsn.5)
rsn.merge <- rsn.5 %>%
  select(-c(LOCATION, TREE_raw, SPECIES_raw, DBH_raw, DA_raw, STATUS, NOTES, OfficeNotes, Noted, ReplaceID, Species.count, Species_Max_Growth, Species_Max_Growth_Error, First_Decimal_Error_High, First_Decimal_Error_Low, Last_Year, Last_Decimal_Error_High, New.Growth, Anomaly, Flag_LastObs_SpeciesSpike, sampled_plots_period, notes_samplingarea, X, DAMAGE_severity_3, DAMAGE_severity_2, DAMAGE_severity_1))%>%
  # RENAME THE PLOT AND SUPLOT FOR EASE
  rename(SUBP = "PLOT",
    PLOT = "SITE")

#str(rsn.merge)
```


# Clean Up CAFI
- there is a damage code with broken top/dead top, etc. 
- first I need to update so that we're just working with teh geo sites that I want to - reduce the dataset to the interior alaska sites
- also want to make sure a site has more than 1 measurement so that can look at change
```{r}
str(cafi)
locs <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/452_CAFI_LOCATION_FUZZED_v20241220.csv")
str(locs)
locs.avg <- locs %>%
  #a weird value that messes up the whole 1098 plots
  filter(OID_ !=290)%>%
  # plot1180 has lat and long backwards on one subplot
  mutate(LAT = ifelse(OID_ == 534, LONG, LAT),
         LONG = ifelse(OID_ == 534, -149.2414022, LONG))%>%
  group_by(PLOT) %>%
  summarize(lat = mean(LAT), long= mean(LONG), n = n()) %>%
  # get rid of a plots that don't have teh right lat/long - not sure what's going on with these
  # check these plots in the data - were they discontinued?
  filter(PLOT %!in% c(1206, 1040, 1026))

locs.interior <- locs.avg %>%
  filter(lat > 62.69353) %>%
  filter(!(lat < 63.15401 & long < -143.7223))

library(leaflet)
leaflet(locs.interior) %>%
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    popup = ~as.character(PLOT)
  )

# looks like there are 78 plots in interior alaska 
# now filter the CAFI dataset to just this 
cafi.2 <- cafi %>%
  filter(PLOT %in% locs.interior$PLOT)

unique(cafi.2$PLOT)

# see if there are sites that only have one measurement in this period
cafi.yr <- cafi.2 %>%
  select(PLOT, INVYR) %>%
  distinct() %>%
  group_by(PLOT) %>%
  summarize(n= n()) %>%
  filter(n < 2)
# 1138 and 1140 and 1142 only have one, going to remove these

cafi.3 <- cafi.2 %>%
  filter(PLOT %!in% cafi.yr$PLOT)
```

# Get rid of the diameters for consistency
- this takes care of the Status code such that is only alive or dead
```{r}
cafi.4 <- cafi.3 %>%
  filter(DIA > 3.8)

```


```{r}
# Want to understand waht the DIACHECK looks like and meanings
# CHECK THE ATTACK TYPE OF THINGS

cafi.checks <- cafi.4 %>%
  group_by(PLOT) %>%
  summarize(n= n(), subs = length(unique(SUBP)))

# clean up pest file
pest.cafi <- pest %>%
  filter(Network == "CAFI")

# Get rid of unnecessary columns, rename some
# Then need to clean up the damage codes
# Then need to add additional binary columns etc that are existent in the dataset
cafi.5 <- cafi.4 %>%
  # Remove some columns off the bat that won't go along with the RSN network
  select(-c(SPCD, AZIMUTH, DIST, PREV_STAT, Pre.Status, RECONCILECD, Reconciled_Meaning, PREVDIA, DIACHECK, DIACHECK_Meaning, PREVHT, ACTUALHT, HT, HTCD, HTCD_Meaning, CRHT, CR, PREV_CCLCD, PREV_CCLCD_Meaning, CULL,ATTACKPRG, ATTACK, HTDMP.Raw, Species.Raw, New.Growth, Anomaly, TREE_ID )) %>%
  # Update the Species
  mutate(SPECIES = case_when(
    Species == "Alaska Paper Birch" ~ "BETNEO",
    Species == "Black Spruce" ~ "PICMAR",
    Species == "Balsam Poplar" ~ "POPBAL",
    Species == "White Spruce" ~ "PICGLA",
    Species == "Aspen" ~ "POPTRE",
    Species == "Tamarack" ~ "LARLAR",
    TRUE ~ Species
  )) %>%
  rename(DBH = "DIA"
  ) %>%
  mutate(Plot.Area.M2 = 405, 
         Plot_Size = "20.12x20.12") %>%
  # fix the status code adn make it dead/alive
  mutate(DA = ifelse(STATUSCD_Meaning == "live", 1, 0), 
         DOWN = ifelse(STANDING_DEAD_CD == 0, 0, 1))%>%
  # drop the old codes
  select(-c(STATUSCD, STATUSCD_Meaning, Species, CCLCD, CCLCD_Meaning,ATTACK_Meaning, ATTACKPRG_Meaning, STANDING_DEAD_CD, STANDING_DEAD_CD_Meaning,DECAYCD,AGENTCD, AGENTCD_Meaning, MORTYR, NOTE, OFFICE, PREV_STAT_Meaning)) %>%
  # Going to work on the Damage stuff now....
  left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD1 == Code)) %>%
  rename(DAMAGE_1 = "Damage") %>%
    select(-c(Network, Common_Name))%>%
    left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD2 == Code)) %>%
  rename(DAMAGE_2 = "Damage") %>%
    select(-c(Network, Common_Name))%>%
    left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD3 == Code)) %>%
  rename(DAMAGE_3 = "Damage") %>%
    select(-c(Network, Common_Name)) %>%
      left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD4 == Code)) %>%
  rename(DAMAGE_4 = "Damage") %>%
    select(-c(Network, Common_Name)) %>%
  # damage severity is already aligned
  # getting rid of hte damage severity meanings
  select(-c(DAMSEV1_Meaning, DAMSEV2_Meaning, DAMSEV3_Meaning, DAMSEV4_Meaning, DAMAGE_AGENT_CD1, DAMAGE_AGENT_CD2, DAMAGE_AGENT_CD3, DAMAGE_AGENT_CD4)) %>%
  # do some more renaming things
  rename(Year = "INVYR") %>%
  mutate(n_plot = ifelse(PLOT %in% c(1078, 1079), 1, 3), 
         sampled_plots = n_plot)
  
  
str(cafi.5)
str(rsn.merge)
```


# Join data
- some qs emerging - how might ht of diameter measurement effect how the data is analyzed 
- have to do 3.8 cm DBH for RSN to be compatible with teh CAFI data - could I include the sapling data here too
- if a tree falls in CAFI - how is that reflected in teh data? the Standing dead, etc.?
- CAFI has mortality year and what likely killed a tree...but I'm going to remove because not totally comparable to RSN

```{r}
# Column names unique to each
setdiff(names(rsn.merge), names(cafi.5))  # in df1 but not df2
setdiff(names(cafi.5), names(rsn.merge))  # in df2 but not df1

# fix the RSN columns
# adding in a DAMSEV4 adn a DAMAGE4
# add in a HTDMP
# if topped, lean, bowed this should be a damage - could add this to damsev4? 

rsn.merge2 <- rsn.merge %>%
  mutate(HTDMP = 4.5, 
         DAMAGE_4 = case_when(
    LEAN == 1 ~ "Damage",
    BOWED == 1 ~ "Damage",
    TOP %in% c("topped", "Topped") ~ "Damage",
    suppressWarnings(as.numeric(TOP)) %>% between(0, 90) ~ "Damage",
    TRUE ~ NA_character_
  ), 
         DAMSEV4 = NA) %>%
  select(-c(DATE, DAMAGE, SEVERITY, TOP, LEAN, BOWED)) %>%
  # update the damaged to include 
  mutate(DAMAGED = as.integer(if_any(c(DAMAGE_1, DAMAGE_2, DAMAGE_3, DAMAGE_4), ~ !is.na(.x))),
         Network = "RSN") %>%
  # must be beyond 2010; gets rid of some of the FP and UP earlier years
  filter(Year > 2010)

# fix the cafi columns
# add a column for yes/no damage (DAMAGED)
# add a max_damage column for greatest number
# make sample.rd from cycle, add sample.yr for consistency - might have ot mess around with this later

# fisx the data ages to get the rough age class
cafi.ages.2 <- cafi.ages %>%
  select(site, StdAge)%>%
  rename(PLOT = "site") %>%
  mutate(age_class_yr = case_when(
    StdAge < 40 ~ "Young", 
    StdAge > 40 & StdAge < 61 ~ "Intermediate", 
    StdAge > 60 ~ "Mature"
  )) %>%
  select(-StdAge)


cafi.merge <- cafi.5 %>%
  mutate(DAMAGED = as.integer(if_any(c(DAMAGE_1, DAMAGE_2, DAMAGE_3, DAMAGE_4), ~ !is.na(.x))), 
         max_damage = pmax(DAMSEV1,DAMSEV2,DAMSEV3, DAMSEV4), 
         sample.rd = CYCLE, 
         sample.yr = Year) %>%
  select(-CYCLE) %>%
  left_join(cafi.ages.2) %>%
  mutate(Network = "CAFI")


# for CAFI, I want the last measurement to be in the 2020s
cafi.2020 <- cafi.merge %>%
  filter(Year > 2020) %>%
  select(PLOT)%>%
  distinct()

cafi.merge2 <- cafi.merge %>%
  filter(PLOT %in% cafi.2020$PLOT)%>%
  # I also want to have more recent years
  filter(Year > 2010)

# merge
merged <- rbind(rsn.merge2, cafi.merge2)

```

# Some visual checks on the merged data
## sampling efforts
- i reckon that this will show more on how we want to break up the RSN network - want the 2020s measurement and the one previous

```{r}
mature <- merged %>%
  filter(age_class_yr == "Mature")
intermediate<- merged %>%
  filter(age_class_yr == "Intermediate")
young <- merged %>%
  filter(age_class_yr == "Young")
tbd <- merged %>%
  filter(is.na(age_class_yr))

ggplot(data = mature, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Mature")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = intermediate, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Intermediate")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = young, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Young")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = tbd, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("TBD")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

```

## Notes 
- ages seems to not be measured in 2025 for the ones in Zach's list!
- still need to get 3.8 dbh for RSN




