---
title: "Merging_RSN_CAFI"
author: "OLH"
date: "2026-05-21"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```


# Read in Data
```{r}
## RSN Datasets
rsn <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/QAQC_Data/Updated/RSN_Inventory_QAQC_Final_2026-04-09.csv')

site.yrs <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/site_years.csv')

## CAFI Datasets
cafi <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/CAFI_Inventory_QAQC_Final.csv")
cafi.ages <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/cafi_loc_deciduous.csv")

## Metadata
pest <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/Data_checks/Pest_merge_check.csv')

```

# RSN Processing

## Data cleaning
- remove sites not in the RSN?
- Get rid of trees that are below 2.5 cm at DBH (Unless they shrink down)
- Get rid of trees that are only ever NA
- Remove one SALBEB obs

```{r}
# sites not in the RSN
non_rsn_sites <- c("FP0A", "FP0B","FP0C", "FP1A","FP1B","FP1C", "FP2A", "FP2B","FP2C",   "FP3A","FP3B","FP3C",'FP4A', 'FP4B', 'FP4C', 'FP5B', "UP1A", "UP1B","UP1C","UP2A", "UP2B", "UP2C", "UP3C", "UP3A", "UP3B")

# initial updates to RSN dataset
rsn.2 <- rsn %>%
  # remove salbeb
  filter(SPECIES != "SALBEB") %>%
  # remove sites not in the RSN
  filter(SITE %!in% non_rsn_sites)%>%
  # set up for removing small trees
  arrange(UniqueID, Year) %>% # Ensure data is in chronological order
  group_by(UniqueID) %>%
  # keep NA DBH, DBH > 2.5, or if the DBH exceeds 2.5, it will keep subsequent observatoins
  # this code above removes 4693 observations
  filter(is.na(DBH) | DBH >= 2.49 | cumany(DBH >= 2.49)) %>%
  ungroup()

# see which trees only have an NA - this is after the seedlings have been removed, so in many cases these had an intiial measurement that was NA and corresponded to somehting like 0.7cm
na.trees <- rsn.2%>%
  group_by(UniqueID) %>%
  filter(all(is.na(DBH))) #%>%
  #distinct(UniqueID)

# remove the na trees only
rsn.3 <- rsn.2 %>%
  filter(UniqueID %!in% na.trees$UniqueID) %>%
  # get rid of sampling rounds before the formal RSN initiation
  filter(Year> 2007)

```

## How many trees start off above 5
- There are a lot of trees over 5cm for first DBH - 624 trees i think and it is mostly FP5A and FP5D (430 is these two)
-I think for FP5A and FP5D, I think that we missed a sampling interval.... so these tries might have existed 
- taking care of that, i think its really 213 that are concerning in terms of over 5; really only over a handful that are over 6 - seem like about 200 trees in total and then only 73 if looking at 6cm

```{r}
early.trees <- rsn.3 %>%
  filter(Year < 2017)

late <- rsn.3 %>%
  filter(UniqueID %!in% early.trees$UniqueID) %>%
  filter(DBH > 5) #%>%
  # select(UniqueID) %>%
  # distinct()

ggplot(late, aes(DBH))+
  geom_histogram()+
  facet_wrap(~SPECIES, scales = "free")

# going to see if these late ones are measured twice?
late2 <- rsn.3 %>%
  filter(UniqueID %in% late$UniqueID) %>%
  filter(DBH < 6)

late3 <- rsn.3 %>%
  filter(UniqueID %in% late$UniqueID) %>%
  filter(UniqueID %!in% late2$UniqueID) %>%
  group_by(SITE) %>%
  summarize(n=n())

# Can run or not if wanting to get rid of these observations; right now its 73 trees that show up greater than 6 in the first obs round
# rsn.4 <- rsn.3 %>%
#   filter(UniqueID %!in% late3$UniqueID)
#   

```

## Clean up the damage and severity columsn - only done in the final sampling period
```{r}
unique(rsn.3$DAMAGE)
unique(rsn.3$LOCATION)
unique(rsn.3$SEVERITY)

# this code was done to identify all of the unique pests that were a thing
# unique.pests <- rsn.3 %>%
#     mutate(DAMAGE2 = case_when(
#     DAMAGE == "Girdled and Black rust" ~ "Girdled, Black rust", 
#     DAMAGE == "Leafroller Leafminer" ~ "Leafroller, Leafminer",
#     TRUE ~ DAMAGE
#   )) %>%
#   select(DAMAGE, DAMAGE2) %>%
#   distinct() %>%
#   mutate(DAMAGE_type = strsplit(DAMAGE, ",\\s*")) %>% 
#   unnest_wider(DAMAGE_type, names_sep = "_")

# prep teh pest dataset for RSN analysis
pests.2 <- pest %>%
  filter(Network == "RSN") %>%
  distinct(Common_Name, .keep_all = TRUE)

rsn.4 <- rsn.3 %>%
  # there are a few categories that need to be fixed so that they separate 
  mutate(DAMAGE = case_when(
    DAMAGE == "Girdled and Black rust" ~ "Girdled, Black rust", 
    DAMAGE == "Leafroller Leafminer" ~ "Leafroller, Leafminer",
    TRUE ~ DAMAGE
  )) %>%
  mutate(DAMAGED = ifelse(DAMAGE == "", 0, 1),
    DAMAGE_type = strsplit(DAMAGE, ",\\s*")) %>% 
  unnest_wider(DAMAGE_type, names_sep = "_")%>%
  # Step 1 - needing to clean up the names based on the pests lookup table
  left_join(pests.2, by = c("DAMAGE_type_1" = "Common_Name")) %>%
  mutate(DAMAGE_1 = Damage) %>%
  select(-c(Network, Code, Damage)) %>%
  left_join(pests.2, by = c("DAMAGE_type_2" = "Common_Name")) %>%
  mutate(DAMAGE_2 = Damage) %>%
  select(-c(Network, Code, Damage)) %>%
  left_join(pests.2, by = c("DAMAGE_type_3" = "Common_Name")) %>%
  mutate(DAMAGE_3 = Damage) %>%
  select(-c(Network, Code, Damage, DAMAGE_type_1, DAMAGE_type_2, DAMAGE_type_3))%>%
  # Now will work on the severity
  mutate(SEVERITY = case_when(
    SEVERITY == "1.1" ~ "1,1",
    TRUE ~ SEVERITY
  )) %>%
  mutate(DAMAGE_severity = strsplit(SEVERITY, ",\\s*")) %>% 
  unnest_wider(DAMAGE_severity, names_sep = "_") %>%
  mutate(DAMAGE_severity_1 = case_when(
    DAMAGE_severity_1 == "4 (36-67%)" ~ "4",
    DAMAGE_severity_1 == "3 (16-35%)" ~ "3",
    DAMAGE_severity_1 == "1 (Trace to 5%)" ~ "1",
    DAMAGE_severity_1 == "5 (68-100%)" ~ "5",
    DAMAGE_severity_1 == "2 (6-15%)"  ~ "2",
    DAMAGE_severity_1 == "very severe" ~ "5",
    DAMAGE_severity_1 == "mo" ~ "3",
    DAMAGE_severity_1 == "mild" ~ "1",
    DAMAGE_severity_1 == "mi" ~ "1",
    DAMAGE_severity_1 == "12" ~ "2",
    TRUE ~ DAMAGE_severity_1
  ))%>%
  # Update damage to match the 1-4 scale of CAFI
  mutate(DAMSEV1= case_when(
    DAMAGE_severity_1 == "3" ~ "2", 
    DAMAGE_severity_1 == "4" ~ "3", 
    DAMAGE_severity_1 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_1
  ), 
  DAMSEV2 = case_when(
    DAMAGE_severity_2 == "3" ~ "2", 
    DAMAGE_severity_2 == "4" ~ "3", 
    DAMAGE_severity_2 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_2
  ), 
  DAMSEV3= case_when(
    DAMAGE_severity_3 == "3" ~ "2", 
    DAMAGE_severity_3 == "4" ~ "3", 
    DAMAGE_severity_3 == "5" ~ "4",
    TRUE ~ DAMAGE_severity_3
  )
  ) %>%
  # Calculate max severity
  mutate(max_damage =pmax(DAMSEV1,DAMSEV2,DAMSEV3)) 
str(rsn.4)
```


## Merge this with the site years
```{r}
rsn.5 <- rsn.4 %>%
  left_join(site.yrs)
```

## Reduce to the necessary columns
- here is what i want: UniqueID, Year, SUBPLOT, Cycle (?), SITE, TREE, SPECIES, DBH, DA, TOP(?), Lean(?), Bowed(?), DOWN (?), Plot_area, Plot_size, no_plots, damage, damage_loc, damage_severity, QAQC_ACTION

```{r}
str(rsn.5)
rsn.merge <- rsn.5 %>%
  select(-c(LOCATION, TREE_raw, SPECIES_raw, DBH_raw, DA_raw, STATUS, NOTES, OfficeNotes, Noted, ReplaceID, Species.count, Species_Max_Growth, Species_Max_Growth_Error, First_Decimal_Error_High, First_Decimal_Error_Low, Last_Year, Last_Decimal_Error_High, New.Growth, Anomaly, Flag_LastObs_SpeciesSpike, sampled_plots_period, notes_samplingarea, X, DAMAGE_severity_3, DAMAGE_severity_2, DAMAGE_severity_1))%>%
  # RENAME THE PLOT AND SUPLOT FOR EASE
  rename(SUBP = "PLOT",
    PLOT = "SITE")

#str(rsn.merge)
```


# Clean Up CAFI
- there is a damage code with broken top/dead top, etc. 
- first I need to update so that we're just working with teh geo sites that I want to - reduce the dataset to the interior alaska sites
- also want to make sure a site has more than 1 measurement so that can look at change
```{r}
str(cafi)
locs <- read.csv("/Users/olhajek/Desktop/RSN/CAFI_data/452_CAFI_LOCATION_FUZZED_v20241220.csv")
str(locs)
locs.avg <- locs %>%
  #a weird value that messes up the whole 1098 plots
  filter(OID_ !=290)%>%
  # plot1180 has lat and long backwards on one subplot
  mutate(LAT = ifelse(OID_ == 534, LONG, LAT),
         LONG = ifelse(OID_ == 534, -149.2414022, LONG))%>%
  group_by(PLOT) %>%
  summarize(lat = mean(LAT), long= mean(LONG), n = n()) %>%
  # get rid of a plots that don't have teh right lat/long - not sure what's going on with these
  # check these plots in the data - were they discontinued?
  filter(PLOT %!in% c(1206, 1040, 1026))

locs.interior <- locs.avg %>%
  filter(lat > 62.69353) %>%
  filter(!(lat < 63.15401 & long < -143.7223))

library(leaflet)
# can also do locs.avg to see what the full map looks like before :) 
leaflet(locs.interior) %>%
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    popup = ~as.character(PLOT)
  )

# looks like there are 78 plots in interior alaska 
# now filter the CAFI dataset to just this 
cafi.2 <- cafi %>%
  filter(PLOT %in% locs.interior$PLOT)

unique(cafi.2$PLOT)

# see if there are sites that only have one measurement in this period
cafi.yr <- cafi.2 %>%
  select(PLOT, INVYR) %>%
  distinct() %>%
  group_by(PLOT) %>%
  summarize(n= n()) %>%
  filter(n < 2)
# 1138 and 1140 and 1142 only have one, going to remove these


cafi.3 <- cafi.2 %>%
  filter(PLOT %!in% cafi.yr$PLOT)
unique(cafi.3$PLOT)
```

# Get rid of the diameters for consistency
- this takes care of the Status code such that is only alive or dead
- this code gets rid of 3 sites in teh CAFI inventory data - 72 sites here
```{r}
cafi.4 <- cafi.3 %>%
  filter(DIA > 3.8)
unique(cafi.4$PLOT)
```


```{r}
# Want to understand waht the DIACHECK looks like and meanings
# CHECK THE ATTACK TYPE OF THINGS

cafi.checks <- cafi.4 %>%
  group_by(PLOT) %>%
  summarize(n= n(), subs = length(unique(SUBP)))

# clean up pest file
pest.cafi <- pest %>%
  filter(Network == "CAFI")

# Get rid of unnecessary columns, rename some
# Then need to clean up the damage codes
# Then need to add additional binary columns etc that are existent in the dataset
cafi.5 <- cafi.4 %>%
  # Remove some columns off the bat that won't go along with the RSN network
  select(-c(SPCD, AZIMUTH, DIST, PREV_STAT, Pre.Status, RECONCILECD, Reconciled_Meaning, PREVDIA, DIACHECK, DIACHECK_Meaning, PREVHT, ACTUALHT, HT, HTCD, HTCD_Meaning, CRHT, CR, PREV_CCLCD, PREV_CCLCD_Meaning, CULL,ATTACKPRG, ATTACK, HTDMP.Raw, Species.Raw, New.Growth, Anomaly, TREE_ID )) %>%
  # Update the Species
  mutate(SPECIES = case_when(
    Species == "Alaska Paper Birch" ~ "BETNEO",
    Species == "Black Spruce" ~ "PICMAR",
    Species == "Balsam Poplar" ~ "POPBAL",
    Species == "White Spruce" ~ "PICGLA",
    Species == "Aspen" ~ "POPTRE",
    Species == "Tamarack" ~ "LARLAR",
    TRUE ~ Species
  )) %>%
  rename(DBH = "DIA"
  ) %>%
  mutate(Plot.Area.M2 = 405, 
         Plot_Size = "20.12x20.12") %>%
  # fix the status code adn make it dead/alive
  mutate(DA = ifelse(STATUSCD_Meaning == "live", 1, 0), 
         DOWN = ifelse(STANDING_DEAD_CD == 0, 0, 1))%>%
  # drop the old codes
  select(-c(STATUSCD, STATUSCD_Meaning, Species, CCLCD, CCLCD_Meaning,ATTACK_Meaning, ATTACKPRG_Meaning, STANDING_DEAD_CD, STANDING_DEAD_CD_Meaning,DECAYCD,AGENTCD, AGENTCD_Meaning, MORTYR, NOTE, OFFICE, PREV_STAT_Meaning)) %>%
  # Going to work on the Damage stuff now....
  left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD1 == Code)) %>%
  rename(DAMAGE_1 = "Damage") %>%
    select(-c(Network, Common_Name))%>%
    left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD2 == Code)) %>%
  rename(DAMAGE_2 = "Damage") %>%
    select(-c(Network, Common_Name))%>%
    left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD3 == Code)) %>%
  rename(DAMAGE_3 = "Damage") %>%
    select(-c(Network, Common_Name)) %>%
      left_join(pest.cafi, by = join_by(DAMAGE_AGENT_CD4 == Code)) %>%
  rename(DAMAGE_4 = "Damage") %>%
    select(-c(Network, Common_Name)) %>%
  # damage severity is already aligned
  # getting rid of hte damage severity meanings
  select(-c(DAMSEV1_Meaning, DAMSEV2_Meaning, DAMSEV3_Meaning, DAMSEV4_Meaning, DAMAGE_AGENT_CD1, DAMAGE_AGENT_CD2, DAMAGE_AGENT_CD3, DAMAGE_AGENT_CD4)) %>%
  # do some more renaming things
  rename(Year = "INVYR") %>%
  mutate(n_plot = ifelse(PLOT %in% c(1078, 1079), 1, 3), 
         sampled_plots = n_plot)
  
unique(cafi.5$PLOT)  
str(cafi.5)
str(rsn.merge)
```


# Join data
- some qs emerging - how might ht of diameter measurement effect how the data is analyzed 
- have to do 3.8 cm DBH for RSN to be compatible with teh CAFI data - could I include the sapling data here too
- if a tree falls in CAFI - how is that reflected in teh data? the Standing dead, etc.?
- CAFI has mortality year and what likely killed a tree...but I'm going to remove because not totally comparable to RSN

```{r}
# Column names unique to each
setdiff(names(rsn.merge), names(cafi.5))  # in df1 but not df2
setdiff(names(cafi.5), names(rsn.merge))  # in df2 but not df1

# fix the RSN columns
# adding in a DAMSEV4 adn a DAMAGE4
# add in a HTDMP
# if topped, lean, bowed this should be a damage - could add this to damsev4? 

rsn.merge2 <- rsn.merge %>%
  mutate(HTDMP = 4.5, 
         DAMAGE_4 = case_when(
    LEAN == 1 ~ "Damage",
    BOWED == 1 ~ "Damage",
    TOP %in% c("topped", "Topped") ~ "Damage",
    suppressWarnings(as.numeric(TOP)) %>% between(0, 90) ~ "Damage",
    TRUE ~ NA_character_
  ), 
         DAMSEV4 = NA) %>%
  select(-c(DATE, DAMAGE, SEVERITY, TOP, LEAN, BOWED)) %>%
  # update the damaged to include 
  mutate(DAMAGED = as.integer(if_any(c(DAMAGE_1, DAMAGE_2, DAMAGE_3, DAMAGE_4), ~ !is.na(.x))),
         Network = "RSN") %>%
  # must be beyond 2010; gets rid of some of the FP and UP earlier years
  filter(Year > 2010)


rsn.merge3 <- rsn.merge2 %>%
  filter(DBH > 3.8)

# fix the cafi columns
# add a column for yes/no damage (DAMAGED)
# add a max_damage column for greatest number
# make sample.rd from cycle, add sample.yr for consistency - might have ot mess around with this later

# fisx the data ages to get the rough age class
cafi.ages.2 <- cafi.ages %>%
  select(site, StdAge)%>%
  rename(PLOT = "site") %>%
  mutate(age_class_yr = case_when(
    StdAge < 40 ~ "Young", 
    StdAge > 40 & StdAge < 61 ~ "Intermediate", 
    StdAge > 60 ~ "Mature"
  )) %>%
  select(-StdAge)


cafi.merge <- cafi.5 %>%
  mutate(DAMAGED = as.integer(if_any(c(DAMAGE_1, DAMAGE_2, DAMAGE_3, DAMAGE_4), ~ !is.na(.x))), 
         max_damage = pmax(DAMSEV1,DAMSEV2,DAMSEV3, DAMSEV4), 
         sample.rd = CYCLE, 
         sample.yr = Year) %>%
  select(-CYCLE) %>%
  left_join(cafi.ages.2) %>%
  mutate(Network = "CAFI")


# for CAFI, I want the last measurement to be in the 2020s
cafi.2020 <- cafi.merge %>%
  filter(Year > 2019) %>%
  select(PLOT)%>%
  distinct()


cafi.merge2 <- cafi.merge %>%
  filter(PLOT %in% cafi.2020$PLOT)%>%
  # I also want to have more recent years
  filter(Year > 2010)

cafi.plots <- cafi.merge2%>%
  select(PLOT)%>%
  distinct

# merge
merged <- rbind(rsn.merge2, cafi.merge2)

# try getting rid of the RSN sites
# importantly this gets rid of 37% of the data in RSN, so this is sad to reduce!!
merged.2 <- merged %>%
  filter(DBH > 3.8)
unique(cafi.merge2$PLOT)
```

# Some visual checks on the merged data
## sampling efforts
- i reckon that this will show more on how we want to break up the RSN network - want the 2020s measurement and the one previous

```{r}
mature <- merged %>%
  filter(age_class_yr == "Mature")
intermediate<- merged %>%
  filter(age_class_yr == "Intermediate")
young <- merged %>%
  filter(age_class_yr == "Young")
tbd <- merged %>%
  filter(is.na(age_class_yr))

ggplot(data = mature, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Mature")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = intermediate, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Intermediate")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = young, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Young")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = tbd, aes(as.factor(PLOT), Year, color = Network)) + 
  geom_point() +
  theme_bw()+
  ggtitle("TBD")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

```

# Visualize the tbd
```{r}

str(tbd)
unique(tbd$PLOT)
tbd2 <- tbd %>%
  mutate(PLOT = as.integer(PLOT))%>%
  left_join(locs.interior) %>%
  select(PLOT, long, lat) %>%
  distinct()

leaflet(tbd2) %>%
  addProviderTiles("Esri.WorldTopoMap") %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    fillOpacity = 0.3,
    popup = ~as.character(PLOT)
  )
```

## Notes 
- ages seems to not be measured in 2025 for the ones in Zach's list!
- if i get rid of trees with 3.8 dbh for RSN, then this gets rid of 42% fo the data!!! - understand the distribution of tree sizes among the rsn sites - how does this vary between young and mature stands
- weird that CAFI merged is only 3684 trees across two sampling periods for 24 sites; I want to look into this more - like how many trees per plot/m2 and how does this compare with some o fhte other mature sites; seems reasonable

# Exploring CAFI tree nos
- why so little and how does that compare with the RSN sites
- look at spread of decid/not in the cafi ages file
- CAFI in general seems to have a little lower tree density? is that a function of smaller trees and more deciduous?


# Plot-level m2 trees
- lose most of the dead trees when increase to over 3.8 DBH requirement in the RSN network; don't see the death of a lot of trees
```{r}
str(merged.2)

subp_trees <- merged.2 %>%
  group_by(Year, PLOT, SUBP, SPECIES, Plot.Area.M2, sample.rd, Network, DA, age_class_yr)%>%
  summarize(n=n(), mean_dbh = mean(DBH, na.rm = TRUE), sd_dbh = sd(DBH,na.rm = TRUE)) %>%
  mutate(no_m2 = n/Plot.Area.M2)

plot_trees <- subp_trees %>%
  ungroup()%>%
  group_by(Year, PLOT, SPECIES, Plot.Area.M2, sample.rd, Network, DA, age_class_yr) %>%
  summarize(no_trees = sum(n), avg_no_m2 = mean(no_m2, na.rm = TRUE), avg_dbh = mean(mean_dbh, na.rm= TRUE), sd_dbh = sd(mean_dbh, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(DA ==1) %>%
  filter(SPECIES != "PICEA")

unique(plot_trees$SPECIES)

no_trees <- plot_trees %>%
  group_by(Year, PLOT, Plot.Area.M2, sample.rd, Network, DA, age_class_yr) %>%
  summarize(no_m2 = sum(avg_no_m2)) %>%
  filter(DA == 1)

ggplot(no_trees, aes(Year, no_m2, color = Network))+
  geom_point(alpha = 0.7)+
  theme_bw() 

ggplot(plot_trees, aes(Year, avg_no_m2, color = Network))+
  geom_point(alpha = 0.7)+
  theme_bw() +
  facet_wrap(~SPECIES)
  
```

# compare canopy classes and locations
- CAFI doesn't necessarily give us a ton
```{r}
# read in the data for the RSN network, join with the CAFI info on decid
rsn.meta <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/2026 Data/RSN_Site/RSN_Master_Site.csv')
str(rsn.meta)
str(cafi.ages)

cafi.canopy <- cafi.ages %>%
  select(c(site, Class, Longitude, Latitude)) %>%
  rename(PLOT = "site") %>%
  mutate(Network = "CAFI")
unique(cafi.canopy$Class)
rsn.canopy <- rsn.meta %>%
  select(sitename, canopy_typ, n_dd, w_dd)%>%
  rename(PLOT = "sitename", 
         Longitude = w_dd, 
         Latitude = n_dd) %>%
  mutate(Class = case_when(
    canopy_typ %in% c("birch", "aspen", "hardwood") ~ "Decid", 
    canopy_typ == "mixed" ~ "Mixed",
    TRUE ~ "Spruce"
  )) %>%
  select(-canopy_typ) %>%
  mutate(Network = "RSN")

canopy <- rbind(cafi.canopy, rsn.canopy)
str(canopy)

# join with the merged dataset
merge.canopy <- merged.2 %>%
  left_join(canopy)

# look at a map
canopy <- merged.2 %>%
  left_join(canopy) %>%
  select(PLOT, Class, Network, Latitude, Longitude)%>%
  distinct()
is.null(canopy)

# Color palette for 3 classes
class_pal <- colorFactor(
  palette = c("#D85A30", "#1D9E75", "#378ADD"),  # coral/Decid, teal/Spruce, blue/Mixed
  levels  = c("Decid", "Spruce", "Mixed")
)

leaflet(data = canopy) %>%
  addProviderTiles("Esri.WorldTopoMap") %>%
  
  # CAFI - black outline
  addCircleMarkers(
    data        = filter(canopy, Network == "CAFI"),
    lng         = ~Longitude,
    lat         = ~Latitude,
    radius      = 7,
    color       = "black",
    weight      = 1.5,
    fillColor   = ~class_pal(Class),
    fillOpacity = 0.9,
    popup = ~paste0(
      "<b>Plot:</b> ", PLOT, "<br>",
      "<b>Class:</b> ", Class, "<br>",
      "<b>Network:</b> ", Network
    )
  ) %>%
  
  # RSN - white outline
  addCircleMarkers(
    data        = filter(canopy, Network == "RSN"),
    lng         = ~Longitude,
    lat         = ~Latitude,
    radius      = 7,
    color       = "white",
    weight      = 1.5,
    fillColor   = ~class_pal(Class),
    fillOpacity = 0.9,
    popup = ~paste0(
      "<b>Plot:</b> ", PLOT, "<br>",
      "<b>Class:</b> ", Class, "<br>",
      "<b>Network:</b> ", Network
    )
  ) %>%
  
  addLegend(
    position = "bottomright",
    pal      = class_pal,
    values   = canopy$Class,
    title    = "Forest class",
    opacity  = 0.9
  )

# look at sampling rounds adn what not
str(merge.canopy)

mature <- merge.canopy %>%
  filter(age_class_yr == "Mature")
intermediate<-  merge.canopy %>%
  filter(age_class_yr == "Intermediate")
young <-  merge.canopy %>%
  filter(age_class_yr == "Young")
tbd <-  merge.canopy %>%
  filter(is.na(age_class_yr))

ggplot(data = mature, aes(as.factor(PLOT), Year, color = Class)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Mature")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = intermediate, aes(as.factor(PLOT), Year, color = Class)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Intermediate")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = young, aes(as.factor(PLOT), Year, color = Class)) + 
  geom_point() +
  theme_bw()+
  ggtitle("Young")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))

ggplot(data = tbd, aes(as.factor(PLOT), Year, color =Class)) + 
  geom_point() +
  theme_bw()+
  ggtitle("TBD")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8))
```


# understand 3.8
- what is the impact of filtering out trees less than 3.8 in RSN - how does this impact vary among the different stand ages
- definitley probably primarily affects young deciduous trees (generally accounts for close to 100% of those trees), but PICMAR is small across age classes and regularly is over 50% of the stand (more consistently probably 25, but it is a big part of the PICMAR community too in mature stand as well). Getting rid of this might matter more so for the PICMAR mature than the mature deciduous stands..? hard to say though because don't have that represented in this dataset. I could look at the sapling data and see what is it looking like
```{r}
tot_num <- rsn.merge2 %>%
  group_by(PLOT, age_class_yr, sample.rd) %>%
  summarize(tot_trees = n())

rsn3.8 <- rsn.merge2 %>%
  mutate(dbh_class = case_when(
    DBH < 3.81 ~ "Small",
    DBH > 10 ~ "Large",
    TRUE ~ "Mid"
  )) %>%
  group_by(PLOT, age_class_yr, sample.rd, SPECIES, dbh_class) %>%
  summarize(n = n()) %>%
  left_join(tot_num)%>%
  mutate(n_prop = n/tot_trees) %>%
  filter(SPECIES %!in% c("PICEA", "LARLAR", "POPBAL"))


number <- ggplot(rsn3.8, aes(sample.rd, n, color = age_class_yr))+
  geom_jitter(alpha = 0.7)+
  theme_bw()+
  facet_wrap(~SPECIES+dbh_class)
number


prop <- ggplot(rsn3.8, aes(sample.rd, n_prop, color = age_class_yr))+
  geom_jitter(alpha = 0.7)+
  theme_bw()+
  facet_wrap(~SPECIES+dbh_class)
prop

```



