---
title: "Preliminary Analysis of RSN-CAFI Data"
author: "OLH"
date: "2026-08-03"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: true
    collapsed: true
    theme: spacelab
    fig_caption: true
    number_sections: false
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.align="center", out.width = "100%")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot, lightbox)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```

# Data Overview 
- Inventory
  - CAFI  - concern is really low tree numbers that even when update RSN to be 3.8, they seem really low. But they are in line with the Trugmann CAFI report. Need to follow up with Helene still!!
  - RSN - have incorporated the 2026 data!! Yay! A little confused on why there is all of the 2018 data for TRM1 if it burned in 2015. Following up on that. Also a few really high DBH still get carried over from teh 2026 data and confirm FP5D is in there. Seems that there are a few that are being read in as a dates form teh file. Raw file has them as date format. 
  - In general, thinking about how to include/adjust for sites that burn during the record. Right now, I'm removing them. 

- Tree Age
  - 5 CAFI sites don't have ages, but everything else is aged well. 
  - Prioritizing tree age with tree rings with large CVs and burn
  - Categorizing stand age. Currently have 40-60 as intermediate, but is that appropriate for mixed/decid stands? Also been using the continuous age too. 
  - there are a few sites with tree cores that suggest very different ages (1088 adn 1022) and no notes; how to resolve. Some of the CAFI have the same patterns where for example, the 12 sites that Melissa cored seem to be much older than those that ANdrew or Isabelle identified

- Other metadata
  - RSN mostly complete, CAFI not. Still thinking about derived sources for some of the metadata. 
  - Still need to do a deeper dive on the soil org depth data and how compiled this is
  
- Climate
  - Still need to do this. 
  
- JFSP Seedling data
  - have this for the different time points
  
- CAFI seedling data - haven't looked at this at all, but looks like it exists. 
  

# 1. Join inventory and stand age data

- Create a category for tree age based on first age measured (young, intermediate, mature)

```{r}
# Read in the data
## Inventory data
trees <- read.csv("/Users/olhajek/Desktop/RSN/RSN_proj/Data/harmonized/RSN_CAFI_merged.csv") %>%
  mutate(DBH = ifelse(Network == "CAFI", DBH *2.54, DBH))

## Tree ages## Tree agesdb_has_table()
ages <- read.csv("/Users/olhajek/Desktop/RSN/RSN_proj/Data/harmonized/continuous_tree_ages.csv") 

# Create a category for stand age based on the first measurement
age.cat <- ages %>%
  # select the first sampling year for each plot - will make the age determination from this
  group_by(PLOT) %>%
  slice_min(Year, n = 1) %>%
  ungroup() %>%
  # calculate an age from year and create a category
  mutate(age.2 = 2026 - age, 
         age_category = case_when(
           age.2 < 40 ~ "Young", 
           age.2 > 60 ~ "Mature",
           is.na(age.2) ~ NA_character_,
           TRUE ~ "Intermediate"
         )) %>%
  select(PLOT, age_category)

# Join the categorical stand age with the ages
ages <- left_join(ages, age.cat)

# Join the trees and the stand age
tree.age <- left_join(trees, ages)

# what trees are missing an age - there are 5 sites, we might want to remove these
na.age <- tree.age %>%
  filter(is.na(age_category)) %>%
  select(PLOT, Year) %>%
  distinct()

```

## Read in lat-long data

```{r}
# RSN
rsn.loc <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/Data/2026 Data/RSN_Site/RSN_latlong.csv') %>%
  rename(PLOT = SITE)

# CAFI
cafi.loc <- read.csv('/Users/olhajek/Desktop/RSN/RSN_proj/Data/2026 Data/CAFI/452_CAFI_LOCATION_FUZZED_v20241220.csv')

cafi.loc2 <- cafi.loc %>%
  #a weird value that messes up the whole 1098 plots
  filter(OID_ !=290)%>%
  # plot1180 has lat and long backwards on one subplot
  mutate(LAT = ifelse(OID_ == 534, LONG, LAT),
         LONG = ifelse(OID_ == 534, -149.2414022, LONG))%>%
  group_by(PLOT) %>%
  summarize(lat = mean(LAT), long= mean(LONG)) %>%
  # get rid of a plots that don't have teh right lat/long - not sure what's going on with these
  # check these plots in the data - were they discontinued?
  filter(PLOT %!in% c(1206, 1040, 1026))

# Join
glimpse(rsn.loc)
glimpse(cafi.loc2)

locs <- rbind(rsn.loc, cafi.loc2)

```


# 2. Calculate trajectory/deciduous index and tree densities

- Based on alive trees
- Compare using 2.5 cm and 3.8 cm cutoffs
- Do any sites shift in their trajectory across time points?

- Notes: 
  - maybe 4 sites where there are a significant number of larch trees and SL1A is the only site with PICEA and it is certainly a large proportion of trees (majority) that are in this category. Larch shouldn't affect trajectory because low number mostly. PICEA would affect 1 trajectory if removed, so will update the PICEA to be PICGLA for now because only PICGLA is identified at SLA1 (it's not a lot...)
  - LARLAR currently being removed for the data but can add in allometry equations from Bond and Lambert 2002. This does require a stand age parameter though. Is there a different equation that is better to use?
  - Using the above DBH allometric equations from H. Alexander. Confirmed the values with appendix 1. 
  - There are 97 trees that are labelled as alive, but don't have a DBH. Majority of these are from the current last measurement period. Keeping these in for a density calculation right now, but not included in the biomass estimates because can't scale up with the biomass.
  -deciduous index used but currently no "Open" type possible - so different from the seedling scaling trajectory 
  - 9 sites change trajectories over time

```{r}
# Summarize site level tree densities by species - trying to see the impact of larlar and picea on the overall biomass/stand level sorts of calculations
tree.species <- trees %>%
  group_by(SPECIES,Year, PLOT) %>%
  summarize(n=n()) %>%
  ungroup()%>%
  pivot_wider(names_from = SPECIES, values_from = n)

# Update the dataset for trajectory classification (PICEA --> PICGLA and remove LARLAR for now)
# also remove the dead trees
trees.traj <- tree.age %>%
  mutate(SPECIES = ifelse(SPECIES == "PICEA", "PICGLA", SPECIES)) %>%
  filter(SPECIES != "LARLAR") %>%
  filter(DA != 0) %>%
  # create a column for area sampled
  mutate(area_sampled = Plot.Area.M2 * sampled_plots_period) %>%
  # removing the trees with NA for DBH
  filter(!is.na(DBH))

#FUNCTIONS FOR TREE ALLOMETRY 
#Above DBH trees used in Alexander et al. 2012, equations from Yarie et al. 2007, function based on species and DBH (cm) measurement; output in g/tree
allom.dbh <- function(x, y) {
  treebiomass <- numeric(length(x)) 
  for (i in seq_along(x)) {
    if (x[i] == "BETNEO") {
      a <- 164.18
      b <- 2.29
    } else if (x[i] == "PICGLA") {
      a <- 96.77
      b <- 2.40
    } else if (x[i] == "PICMAR") {
      a <- 271.46
      b <- 1.84
    } else if (x[i] == "POPTRE") {
      a <- 134.10
      b <- 2.26
    } else if (x[i] == "POPBAL") {
      a <- 133.71
      b <- 2.29
    } else {
      a <- NA
      b <- NA
    }
    treebiomass[i] <- a * ((y[i])^b)
  }
  return(treebiomass) 
}

#Calculated by getting a mean biomass for above and below for each species, multiply by the site-level number of below and above trees, add together the biomass of below and above dbh trees of each species then divided by the area sampled to get the total biomass per msq of each vegetation type 
trajectories <- trees.traj %>%
    # removing any super large DBHS!!
  filter(DBH < 35 | is.na(DBH)) %>%
  #Use allometric equation to calculate biomass of individuals measured, both above and below dbh
  mutate(above_dbh_biomass = allom.dbh(SPECIES, DBH)) %>%
  group_by(Network, PLOT, SPECIES, sample.pd, sampled_plots_period, area_sampled) %>%
  #for each species within each site- get the mean value for biomass for above dbh individuals and below dbh individuals. Add all below dbh individuals and all above dbh individuals for each species in each site
  summarise(mean_above_bio = mean(above_dbh_biomass, na.rm = TRUE), count_above = n()) %>%
  ungroup()%>%
  #multiply mean biomass * number of individuals for above and below individuals of each species, get total # of each species at the site level 
  mutate(total_above_bio = mean_above_bio * count_above, total_count = count_above) %>%
  #Get total biomass of each species at each site by getting sum of above and below biomass estimates
  mutate(total_biomass = total_above_bio) %>%
  #set spruce vs. deciduous designations by species
  mutate(veg_type = ifelse(SPECIES %in% c("BETNEO", "POPTRE", "POPBAL"), "decid", "spruce")) %>%
  group_by(Network, PLOT, sample.pd, sampled_plots_period, area_sampled, veg_type) %>%
  #get total spruce and deciduous density and biomass for each site
  summarise(site_count = sum(total_count), site_biomass = sum(total_biomass)) %>%
  #Divide the total count and biomass estimated for site by the area surveyed to get the meter squared density and biomass of each vegetation type
  mutate(dens_msq = site_count/area_sampled) %>%
  mutate(biomass_msq = site_biomass/area_sampled) %>%
  #get rid of columns you don't need 
  select(-c(site_count, site_biomass)) %>%
  #pivot data so you have a column for biomass and density for both spruce and deciduous 
  pivot_wider(names_from = veg_type, values_from = c(dens_msq, biomass_msq)) %>%
  mutate(across(starts_with("dens_msq_") | starts_with("biomass_msq_"), ~replace_na(., 0))) %>%
  #Calculate relative deciduous density, relative deciduous biomass, and deciduous index
  mutate(rel_dens_decid = dens_msq_decid/(dens_msq_decid + dens_msq_spruce), 
         rel_biomass_decid = biomass_msq_decid/(biomass_msq_decid + biomass_msq_spruce), 
         decid.index.msq = (rel_dens_decid + rel_biomass_decid)/2) %>%
  #Categorize sites based on their deciduous index value!
  mutate(canopy= case_when(
    decid.index.msq > 0.66 ~ "decid",
    decid.index.msq < 0.33 ~ "spruce",
    TRUE ~ "mixed"
  ))

# Do any trajectories change over time?
traj.change <- trajectories %>%
  ungroup()%>%
  select(PLOT, canopy) %>%
  unique() %>%
  group_by(PLOT)%>%
  filter(n()>1)

# Join trajectories - want a column that is for the final trajectory
# select just the final trajectory
traj.join <- trajectories %>%
  ungroup()%>%
  filter(!is.na(sampled_plots_period)) %>%
  group_by(PLOT) %>%
  slice_max(sample.pd, n = 1) %>%
  ungroup() %>%
  select(PLOT, canopy)

## THIS IS THE DATASET THAT WE WILL USE MOVING FORWARD
tree.age.traj <- tree.age %>%
  left_join(traj.join) %>%
  # create a column for the tree biomass too
  mutate(above_dbh_biomass = allom.dbh(SPECIES, DBH)) %>%
  # removing any super large DBHS!!
  filter(DBH < 35 | is.na(DBH))

# visualize changes in trajectory over time
traj <- left_join(trajectories, ages)

ggplot(traj, aes(sample.pd, decid.index.msq, group=PLOT))+
  geom_point()+
  geom_line()+
  theme_bw()+ggtitle("DEciduous Values Over Time") +
  geom_hline(yintercept=0.66, linetype= "dashed", color = "red") + 
  geom_hline(yintercept=0.33, linetype= "dashed", color="green") + 
  facet_wrap(~age_category)

# Check how this changes if we just do 3.8 DBH for the RSN sites
trees.traj.3.8 <- trees.traj %>%
  filter(DBH > 3.7)


trajectories3.8 <- trees.traj.3.8 %>%
    # removing any super large DBHS!!
  filter(DBH < 35 | is.na(DBH)) %>%
  #Use allometric equation to calculate biomass of individuals measured, both above and below dbh
  mutate(above_dbh_biomass = allom.dbh(SPECIES, DBH)) %>%
  group_by(Network, PLOT, SPECIES, sample.pd, sampled_plots_period, area_sampled) %>%
  #for each species within each site- get the mean value for biomass for above dbh individuals and below dbh individuals. Add all below dbh individuals and all above dbh individuals for each species in each site
  summarise(mean_above_bio = mean(above_dbh_biomass, na.rm = TRUE), count_above = n()) %>%
  ungroup()%>%
  #multiply mean biomass * number of individuals for above and below individuals of each species, get total # of each species at the site level 
  mutate(total_above_bio = mean_above_bio * count_above, total_count = count_above) %>%
  #Get total biomass of each species at each site by getting sum of above and below biomass estimates
  mutate(total_biomass = total_above_bio) %>%
  #set spruce vs. deciduous designations by species
  mutate(veg_type = ifelse(SPECIES %in% c("BETNEO", "POPTRE", "POPBAL"), "decid", "spruce")) %>%
  group_by(Network, PLOT, sample.pd, sampled_plots_period, area_sampled, veg_type) %>%
  #get total spruce and deciduous density and biomass for each site
  summarise(site_count = sum(total_count), site_biomass = sum(total_biomass)) %>%
  #Divide the total count and biomass estimated for site by the area surveyed to get the meter squared density and biomass of each vegetation type
  mutate(dens_msq = site_count/area_sampled) %>%
  mutate(biomass_msq = site_biomass/area_sampled) %>%
  #get rid of columns you don't need 
  select(-c(site_count, site_biomass)) %>%
  #pivot data so you have a column for biomass and density for both spruce and deciduous 
  pivot_wider(names_from = veg_type, values_from = c(dens_msq, biomass_msq)) %>%
  mutate(across(starts_with("dens_msq_") | starts_with("biomass_msq_"), ~replace_na(., 0))) %>%
  #Calculate relative deciduous density, relative deciduous biomass, and deciduous index
  mutate(rel_dens_decid = dens_msq_decid/(dens_msq_decid + dens_msq_spruce), 
         rel_biomass_decid = biomass_msq_decid/(biomass_msq_decid + biomass_msq_spruce), 
         decid.index.msq3.8 = (rel_dens_decid + rel_biomass_decid)/2) %>%
  #Categorize sites based on their deciduous index value!
  mutate(canopy3.8= case_when(
    decid.index.msq3.8  > 0.66 ~ "decid",
    decid.index.msq3.8  < 0.33 ~ "spruce",
    TRUE ~ "mixed"
  )) %>%
  select(PLOT, sample.pd, decid.index.msq3.8, canopy3.8)

# check the difference
traj.2 <- left_join(traj, trajectories3.8) %>%
  filter(Network == "RSN")

# Do lose several sites having trees if move to 3.8
# generally seem to see decent alignment between trajectories, but tendency to see more deciduous with the 3.8 in intermediate stands
ggplot(traj.2, aes(decid.index.msq,decid.index.msq3.8))+
  geom_point()+
  theme_bw()+
  ggtitle("DEciduous Values - Trees > 3.8cm") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50")+
  geom_hline(yintercept=0.66, linetype= "dashed", color = "red") + 
  geom_hline(yintercept=0.33, linetype= "dashed", color="green") + 
  geom_vline(xintercept=0.66, linetype= "dashed", color = "red") + 
  geom_vline(xintercept=0.33, linetype= "dashed", color="green") + 
  facet_wrap(~age_category)

```


# 3. Summary of the data

- Site sampling lengths by age and trajectory
  - Sites with inventories before 2013 (in particular FP5A, FP5D)
- Stand age summaries by trajectory
  - Stand age: mean, max, min; burn year vs. tree ring — anything to consider here?
  - Cohort vs. mixed age: stand-replacing fire suggests cohort structure, but evidence exists for mixed-age structure too. Does our age + CV data show two structures?
  - Deciduous trees peak earlier than black spruce — how should that reshape the current young/intermediate/mature classifications, which were built around black spruce?
- Tree density comparisons using the 2.5 cm vs. 3.8 cm cutoffs for RSN — how does that change the number of stems?
- maybe a nice map of all of the sites colored by trajectorie and shaped by age or something likethat

- Notes 
  - there are 263 trees that are dead (DA = 0) but have DBH, most of these are from CAFI. Look into why this might be the case!
  
## Data length
```{r}
glimpse(tree.age.traj)

# there are a number of "dead" trees with measured DBH primarily from the CAFI network
check <- tree.age.traj %>%
  filter(DA == 0)%>%
  filter(!is.na(DBH))

# create a summary of site with number of trees, dead and alive and then the change
data.summary <- tree.age.traj %>%
  group_by(Network, sample.pd, age_category, canopy, PLOT, DA)%>%
  summarize(no_trees = n(), biomass = sum(above_dbh_biomass, na.rm = T)) %>%
  pivot_wider(
    names_from = DA,
    values_from = c(no_trees, biomass),
    names_sep = "_"
  )

glimpse(data.summary)
ggplot(data = subset(data.summary, data.summary$age_category == "Mature"), aes(sample.pd, PLOT, color = canopy, shape = Network))+
  geom_point()+ ggtitle("Mature")+
  theme_bw()+
  facet_wrap(~age_category, scales= "free")

ggplot(data = subset(data.summary, data.summary$age_category == "Intermediate"), aes(sample.pd, PLOT, color = canopy, shape = Network))+
  geom_point()+
  ggtitle("Intermediate")+
  theme_bw()+
  facet_wrap(~age_category, scales= "free")

ggplot(data = subset(data.summary, data.summary$age_category == "Young"), aes(sample.pd, PLOT, color = canopy, shape = Network))+
  geom_point()+ ggtitle("Young")+
  theme_bw()+
  facet_wrap(~age_category, scales= "free")

```


## Stand age
- just focusing on RSN to look at these dynamics BUT could do them all with the RSN and CAFI merged sites...will do 
- what is cbsnbc as a SPECIES?
- some pretty substantial variation amont ages for a large number of mature spruce sites in particular
```{r}
rsn.ages <- read.csv("/Users/olhajek/Desktop/RSN/RSN_proj/Data/Tree_Age_Data/Compiled/RSN_TreeAge_Compiled.csv") %>%
  rename(PLOT = Site) %>%
  left_join(traj.join) %>%
  # fix species
  mutate(SPECIES = case_when(
    Species == "LALA" ~ "LARLAR", 
    Species == "LALA" ~ "LARLAR",
    Species == "PICLGLA" ~ "PICGLA",
    Species == "Picmar" ~ "PICMAR",
    Species == "PIGL" ~ "PICGLA",
    Species == "PIMA" ~ "PICMAR",
    TRUE ~ Species
  ))

glimpse(rsn.ages)

ggplot(rsn.ages, aes(Pith_Year, PLOT, color = SPECIES))+
  geom_point()+
  theme_bw() +
  facet_wrap(~canopy, scales = "free")

# subset to zoom in 
ages.range <- ages %>%
  filter(max_ty-min_ty > 50)

rsn.ages2 <- rsn.ages %>%
  filter(PLOT %in% ages.range$PLOT)

ggplot(data = rsn.ages2, aes(Pith_Year, PLOT, color = SPECIES))+
  geom_point()+
  ggtitle("Plots with a range greater than 50yrs")+
  theme_bw() +
  xlim(1775, 2000)

```


## Tree density & biomass
- Why is it that CAFI seems to have SUCH fewer trees. Even if i go down to 3.8cm then it still feels really low. 
- biomass values seem in line with H.Alexander ecosphere but density seems low. likely because of 2.5 cm DBH req. - using 3.8cm cutoff really doesn't make that large of a difference for biomass, even though it really reduces the number of trees so most of biomass in larger trees vs. many
```{r}
data.summary.tree <- tree.age.traj %>%
  mutate(sample.area = sampled_plots_period * Plot.Area.M2)%>%
  group_by(Network, sample.pd, age_category, canopy, PLOT, DA, SPECIES, sample.area)%>%
  summarize(no_trees = n(), biomass = sum(above_dbh_biomass, na.rm = T)) %>%
  mutate(no_trees_m2 = no_trees/sample.area, biomass_m2 = biomass/sample.area)

# Density
glimpse(data.summary.tree)

# Just live trees for the final sampling period
dst.2 <- data.summary.tree %>%
  filter(DA == 1) %>%
  filter(sample.pd == 3)  %>%
  filter(!is.na(age_category))
ggplot(data = dst.2,
       aes(x = no_trees_m2, y = PLOT, fill = SPECIES)) +
  geom_col() +
  theme_bw() +
  labs(x = "Number of Trees (m2)", y = "PLOT", fill = "Species") +
  facet_wrap(~age_category, scales = "free_y") +
  ggtitle("Alive Trees in Last Sampling")
  
# Biomass
ggplot(data = dst.2,
       aes(x = biomass_m2, y = PLOT, fill = SPECIES)) +
  geom_col() +
  theme_bw() +
  labs(x = "Biomass(g/m2)", y = "PLOT", fill = "Species") +
  facet_wrap(~age_category, scales = "free_y", nrow = 3) +
  ggtitle("Alive Trees in Last Sampling")


# Just deadtrees for the final sampling period
dst.0 <- data.summary.tree %>%
  filter(DA == 0) %>%
  filter(sample.pd == 3) %>%
  filter(!is.na(age_category))
ggplot(data = dst.0,
       aes(x = no_trees_m2, y = PLOT, fill = SPECIES)) +
  geom_col() +
  theme_bw() +
  labs(x = "Number of Trees/m2", y = "PLOT", fill = "Species") +
  facet_wrap(~age_category, scales = "free")+
  ggtitle("Dead Trees in Last Sampling")

# Make a comparable table
data.summary.tree2 <- tree.age.traj %>%
  group_by(Network, sample.pd, age_category, canopy, PLOT, DA, SPECIES) %>%
  summarize(no_trees = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = c(DA, SPECIES),
    values_from = no_trees,
    values_fill = 0
  ) %>%
  arrange(PLOT, sample.pd)



# how would this change with the 3.8 
data.summary.tree2 <- tree.age.traj %>%
  mutate(sample.area = sampled_plots_period * Plot.Area.M2)%>%
  filter (DBH > 3.7) %>%
  group_by(Network, sample.pd, age_category, canopy, PLOT, DA, SPECIES, sample.area)%>%
  summarize(no_trees = n(), biomass = sum(above_dbh_biomass, na.rm = T)) %>%
  mutate(no_trees_m2 = no_trees/sample.area, biomass_m2 = biomass/sample.area)

dst.38 <- data.summary.tree2 %>%
  filter(DA == 1) %>%
  filter(sample.pd == 3)  %>%
  filter(!is.na(age_category))
ggplot(data = dst.38,
       aes(x = no_trees_m2, y = PLOT, fill = SPECIES)) +
  geom_col() +
  theme_bw() +
  labs(x = "Number of Trees/m2", y = "PLOT", fill = "Species") +
  facet_wrap(~age_category, scales = "free_y") +
  ggtitle("Alive Trees in Last Sampling - DBH > 3.8")
  
ggplot(data = dst.38,
       aes(x = biomass_m2, y = PLOT, fill = SPECIES)) +
  geom_col() +
  theme_bw() +
  labs(x = "Biomass (g/m2)", y = "PLOT", fill = "Species") +
  facet_wrap(~age_category, scales = "free_y") +
  ggtitle("Alive Trees in Last Sampling - DBH > 3.8")



```

## CAFI Deep Dive
```{r}
# relationship between number of trees and biomass by network and age
ggplot(data = subset(dst.2, dst.2$Network == "CAFI"), aes(no_trees_m2, biomass_m2, color = Network))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age_category)

cafi <- tree.age.traj %>%
  filter(Network == "CAFI") %>%
  group_by(Network, sample.pd, age_category, canopy, PLOT, DA, SUBP, SPECIES, Plot.Area.M2)%>%
  summarize(no_trees = n(), biomass = sum(above_dbh_biomass, na.rm = T)) %>%
  mutate(no_trees_m2 = no_trees/Plot.Area.M2, biomass_m2 = biomass/Plot.Area.M2)%>%
  filter(sample.pd == 3)


ggplot(cafi, aes(no_trees_m2, biomass_m2, color = canopy))+
  geom_point()+
  theme_bw()+
  facet_wrap(~age_category)

```


## Map

```{r}
locs.2 <- left_join(traj.join, locs)
locs.3 <- left_join(locs.2, ages)

library(leaflet)
library(dplyr)

# --- 1. Define your color and shape mappings ---
# Adjust these names to match your actual trajectory / age_category values
trajectory_colors <- c(
  "spruce"    = "#2c7fb8",
  "decid" = "#d95f0e",
  "mixed"      = "#31a354"
)

age_shapes <- c(
  "Young" = "circle",
  "Intermediate"   = "square",
  "Mature"   = "triangle"
)

# --- 2. Function to build a small SVG icon for a given shape + color ---
make_svg_icon <- function(shape, color, size = 22) {
  shape_svg <- switch(shape,
    "circle"   = sprintf('<circle cx="%1$s" cy="%1$s" r="%2$s" fill="%3$s" stroke="black" stroke-width="1"/>',
                          size / 2, size / 2 - 2, color),
    "square"   = sprintf('<rect x="2" y="2" width="%1$s" height="%1$s" fill="%2$s" stroke="black" stroke-width="1"/>',
                          size - 4, color),
    "triangle" = sprintf('<polygon points="%1$s,2 %2$s,%3$s 2,%3$s" fill="%4$s" stroke="black" stroke-width="1"/>',
                          size / 2, size - 2, size - 2, color)
  )
  svg <- sprintf('<svg xmlns="http://www.w3.org/2000/svg" width="%1$s" height="%1$s">%2$s</svg>', size, shape_svg)
  paste0("data:image/svg+xml;utf8,", utils::URLencode(svg, reserved = TRUE))
}

# --- 3. Attach an icon URL to every row of your site data ---
# site.data needs: lat, lon, trajectory, age_category (and anything you want in popups)
site.data <- locs.3 %>%
  mutate(icon_url = mapply(make_svg_icon,
                            shape = age_shapes[age_category],
                            color = trajectory_colors[canopy]))

site_icons <- icons(
  iconUrl   = site.data$icon_url,
  iconWidth = 22, iconHeight = 22
)

# --- 4. Build the map ---
leaflet(site.data) %>%
  addProviderTiles(providers$Esri.WorldTopoMap) %>%
  addMarkers(
    lng = ~long, lat = ~lat, icon = site_icons,
    popup = ~paste0("<b>Site:</b> ", PLOT,
                     "<br><b>Trajectory:</b> ", canopy,
                     "<br><b>Age category:</b> ", age_category)
  ) %>%
  addLegend("bottomright",
            colors = trajectory_colors, labels = names(trajectory_colors),
            title = "Trajectory") %>%
  addControl(
    html = paste0(
      "<div style='background:white;padding:6px 8px;border-radius:4px;font-size:12px;'>",
      "<b>Age category</b><br>",
      paste0("<img src='", sapply(names(age_shapes), function(a) make_svg_icon(age_shapes[[a]], "black")),
             "' width='14' height='14' style='vertical-align:middle;'> ", names(age_shapes),
             collapse = "<br>"),
      "</div>"
    ),
    position = "bottomright"
  )
```


# 4. Individual tree growth and mortality

- How does growth vary across stand age and among tree species? Specifically, how does the slope of the yield curve change by species and age?
  - Yield curve: RGR (or g/cm/tree per year) on the y-axis vs. age (years since fire) on the x-axis, faceted/colored by species. Read off slope changes by species and age class.
  - Per-unit biomass vs. RGR — the basic forestry graph, plus follow-on basic forestry diagnostics as exploratory analyses.
  - Same growth-age plot at the site level: each point is a site, x = time since fire, y = g C m^-2^ yr^-1^, with a fit line per tree species x age class to compare slopes. Alternative version facets by tree species with one line per canopy/stand type.

##Individual metric calculations -periods 1 to 3
```{r}
# Create tree level metrics
# RGR  - biomass, DBH, BA
tree.metrics <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    # BA in m2
    BA = 0.00007854 * (DBH)^2,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    RBA = (log(BA) - log(lag(BA)))/Year_diff,
    Year_before = lag(Year), 
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    biomass_RGR_XW = ((above_dbh_biomass - lag(above_dbh_biomass))/lag(above_dbh_biomass))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_"))

ggplot(tree.metrics, aes(above_dbh_biomass,fill = age_category))+
  geom_histogram(alpha=0.5)+
  facet_wrap(~SPECIES, scales = "free")
```

## Individ Graphs

- How does growth vary across stand age and among tree species? Specifically, how does the slope of the yield curve change by species and age?
  - Yield curve: RGR (or g/cm/tree per year) on the y-axis vs. age (years since fire) on the x-axis, faceted/colored by species. Read off slope changes by species and age class.
  - Per-unit biomass vs. RGR — the basic forestry graph, plus follow-on basic forestry diagnostics as exploratory analyses.
  - Same growth-age plot at the site level: each point is a site, x = time since fire, y = g m^-2 yr^-1, with a fit line per tree species x age class to compare slopes. Alternative version facets by tree species with one line per canopy/stand type.
  
  
### Notes
- I think still some issues wiht the difference - these seem like decimal issues; why are these happening. listed two trees that are substantial
GRI4_14_1728
MDM1_1_28

- these seem really high - opposite issue
TRM4_6_231

but there are some others that dont seem right...

### Yield Curve
```{r}
# across all trees
tree.metrics2 <- tree.metrics %>%
  filter(!is.na(biomass_change)) %>%
  filter(SPECIES != "POPBAL")

ggplot(tree.metrics2, aes(age2, biomass_change, color = canopy))+
  geom_point()+
  theme_bw()+
  facet_wrap(~SPECIES, scales = "free")

# at the site level
glimpse(tree.metrics2)
# look at high 
high.low <- tree.metrics2 %>%
  filter(biomass_change > 900) %>%
  select(PLOT, biomass_change, DBH, lag_DBH, Year_diff, UniqueID, SPECIES)

tree.metrics3 <- tree.metrics2 %>%
  filter(Annual_Growth < 2) %>%
  filter(Annual_Growth > -0.5) %>%
  group_by(Network, PLOT, sample.pd, age2, SPECIES, canopy, Year_diff) %>%
  summarize(mean_biomass_change = mean(biomass_change), RGR_change = mean(biomass_RGR), RGR_changeXW = mean(biomass_RGR_XW), RGR_2 = mean(RGR)) %>%
  ungroup()%>%
  mutate(anpp = mean_biomass_change/Year_diff, RGR_yr_XW = RGR_changeXW /Year_diff,
         RGR_yr = RGR_change/Year_diff)

ggplot(tree.metrics3, aes(age2, mean_biomass_change, color = canopy, shape = Network))+
  geom_point()+
  theme_bw()+
  ylab("Growth g/m2/yr") +
  ggtitle("Site-level Site-level Average Species Growth (biomass)")+
  facet_wrap(~SPECIES, scales = "free")
  
ggplot(tree.metrics3, aes(age2, RGR_change, color = canopy, shape = Network))+
  geom_point()+
  theme_bw()+
  ggtitle("Site-level average RGR by species")+
  facet_wrap(~SPECIES, scales = "free")

ggplot(tree.metrics3, aes(age2, RGR_changeXW, color = canopy, shape = Network))+
  geom_point()+
  theme_bw()+
  ggtitle("Site-level average RGR change by species using XW Calc")+
  facet_wrap(~SPECIES, scales = "free")

```

### Yield vs. BA/Densities
- biomass vs. density
- RGR vs. biomass
- RGR vs. density
- RGR vs. BA
- BA vs. density
- biomass vs. BA
- looking at these in the context of age, species, dead/alive, canopy type - maybe can see the overyielding?
- want a dataframe that has total biomass/density/ba at stand level for dead and alive - all trees combined one that has these features for each species (maybe only alive at the species level?)
- going to focus on Biomass change vs. other metrics. and maybe will start with the 1-3 sample periods?


- Notes: there are a lot of sites with just a few trees of a species (i.e., 1-3), might make sense to filter these out
```{r}
# live tree 
tree.metrics.sp <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    BA = 0.00007854 * (DBH)^2,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    Year_before = lag(Year), 
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3, this will be so we can get all of the BA and biomass of the live trees by species
  filter(sample.pd == 3) %>%
  filter(SPECIES != "POPBAL") %>%
# get rid of the crazy high stuff
  filter(
  is.na(Annual_Growth) | 
    (Annual_Growth < 2 & Annual_Growth > -0.5)) %>%
  group_by(Network, PLOT, sample.pd, age2, SPECIES, canopy, sample.area) %>%
  summarize(
    # number of trees 
    n_trees = n()/first(sample.area), 
    # BA and biomass
    BA_sp = sum(BA)/first(sample.area), biomass_sp = sum(above_dbh_biomass)/first(sample.area),
    # mean Growth rates
    mean_biomass = mean(biomass_change, na.rm = T), mean_RGR = mean(biomass_RGR, na.rm = T))


# stand metrics 
stand <- tree.age.traj %>%
  filter(sample.pd == 3) %>%
  # calculate BA for each tree
  mutate(
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    BA = 0.00007854 * (DBH)^2,
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  group_by(Network, PLOT, DA, sample.pd, age, age_category,canopy, sample.area) %>%
  summarize(
    # number of trees 
    n_trees = n()/first(sample.area), 
    # BA and biomass
    BA = sum(BA, na.rm = T)/first(sample.area), biomass = sum(above_dbh_biomass, na.rm = T)/first(sample.area)) %>%
  pivot_wider(names_from = DA,
              values_from = c(n_trees, BA, biomass)) %>%
  select(-c(biomass_0, BA_0)) %>%
  mutate(n_trees_0 = replace_na(n_trees_0, 0))


# join the two dataframes
stand.sp <- left_join(tree.metrics.sp, stand)

# graphs!
glimpse(stand.sp)
# - biomass vs. density 
ggplot(stand, aes(n_trees_1, BA_1, color = canopy))+
  geom_point()+
  xlab("No. Trees/m2")+
  ylab("Basal Area/m2")+
  theme_bw()+
  facet_wrap(~age_category)

ggplot(stand.sp, aes(n_trees_1, biomass_sp, color = SPECIES, shape = Network))+
  geom_point()+
  #geom_line(aes(group=PLOT))+
  xlab("No. Trees")+
  ylab("Species Biomass")+
  theme_bw()+
  facet_wrap(~age_category)

# - RGR vs. biomass
stand.sp2 <- stand.sp %>%
  filter(SPECIES %!in% c("LARLAR", "PICEA"))

# this plot doesn't hav emany of the young sites b/c so many didn't initially have trees
ggplot(stand.sp2, aes(biomass_1, mean_RGR, color = age_category, shape = canopy))+
  geom_point()+
  #geom_line(aes(group=PLOT))+
  xlab("Stand Biomass m2")+
  ylab("Species Biomass RGR")+
  theme_bw()+
  facet_wrap(~SPECIES)

# - RGR vs. density

ggplot(stand.sp2, aes(n_trees_1, mean_RGR, color = age_category, shape = canopy))+
  geom_point()+
  #geom_line(aes(group=PLOT))+
  xlab("No. Trees/m2")+
  ylab("Species Biomass RGR")+
  theme_bw()+
  facet_wrap(~SPECIES, scales = "free")

ggplot(stand.sp2, aes(n_trees, mean_RGR, color = age_category, shape = canopy))+
  geom_point()+
  #geom_line(aes(group=PLOT))+
  xlab("No. Trees of same species/m2")+
  ylab("Species Biomass RGR")+
  theme_bw()+
  facet_wrap(~SPECIES, scales = "free")

# - RGR vs. BA

ggplot(stand.sp2, aes(BA_1, mean_RGR, color = age_category, shape = canopy))+
  geom_point()+
  #geom_line(aes(group=PLOT))+
  xlab("BA/m2")+
  ylab("Species Biomass RGR")+
  theme_bw()+
  facet_wrap(~SPECIES)

```

##Individual metric calculations -periods 1, 2, and 3 

```{r}

tree.metrics.1.2 <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    # BA in m2
    BA = 0.00007854 * (DBH)^2,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    RBA = (log(BA) - log(lag(BA)))/Year_diff,
    Year_before = lag(Year), 
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    biomass_RGR_XW = ((above_dbh_biomass - lag(above_dbh_biomass))/lag(above_dbh_biomass))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_"))

tree.metrics.2.3 <- tree.age.traj %>%
  filter(sample.pd == 2 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    # BA in m2
    BA = 0.00007854 * (DBH)^2,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    RBA = (log(BA) - log(lag(BA)))/Year_diff,
    Year_before = lag(Year), 
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    biomass_RGR_XW = ((above_dbh_biomass - lag(above_dbh_biomass))/lag(above_dbh_biomass))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_"))

tree.metrics.1.2.summary <- tree.metrics.1.2%>%
  filter(Annual_Growth < 2) %>%
  filter(Annual_Growth > 0) %>%
    filter(biomass_change < 6000) %>%
  filter(biomass_change > -100)%>%
  group_by(Network, PLOT, sample.pd, age2, SPECIES, canopy, Year_diff) %>%
  summarize(mean_biomass_change = mean(biomass_change), RGR_change = mean(biomass_RGR), RGR_changeXW = mean(biomass_RGR_XW), RGR_2 = mean(RGR), mean_DBH = mean(DBH), n = n()) %>%
  ungroup()

tree.metrics.2.3.summary <- tree.metrics.2.3%>%
  filter(Annual_Growth < 2) %>%
  filter(Annual_Growth > -0) %>%
  filter(biomass_change < 6000) %>%
  filter(biomass_change > -100)%>%
  group_by(Network, PLOT, sample.pd, age2, SPECIES, canopy, Year_diff) %>%
  summarize(mean_biomass_change = mean(biomass_change), RGR_change = mean(biomass_RGR), RGR_changeXW = mean(biomass_RGR_XW), RGR_2 = mean(RGR), mean_DBH = mean(DBH), n = n()) %>%
  ungroup()

tree.metrics.1.2.3 <- rbind(tree.metrics.1.2.summary, tree.metrics.2.3.summary) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) 

ggplot(tree.metrics.1.2.3, aes(age2, mean_biomass_change, color = canopy, shape = as.factor(sample.pd)))+
  geom_point()+
  theme_bw()+
  ylab("Tree growth - g/tree/yr") +
  ggtitle("Site-level Average Species growth")+
  facet_wrap(~SPECIES)

ggplot(tree.metrics.1.2.3, aes(age2, mean_DBH, color = log10(n), shape = as.factor(sample.pd)))+
  geom_point()+
  scale_color_viridis_c()+
  theme_bw()+
  ylab("Mean DBH (cm)") +
  ggtitle("Site-level Average Species Tree growth")+
  facet_wrap(~SPECIES)

```


# 5. Stand demographics

- How do birth (recruitment) and mortality, alongside individual growth, contribute to stand-level growth and biomass accumulation?
- Relate demographic processes to carbon stocks — outputs in g C m^-2^ yr^-1^. Death-carbon enters as a negative.
- How do these demographic contributions change through succession?
- Is competition or facilitation driving these dynamics? Consider asymmetric competition between growth trajectory and death.
- How do pests and pathogens alter these patterns?
- Bar graphs of absolute carbon-pool size split by demographic driver (growth, birth, death).
- Continuous-age version of the same: x = age since fire, y = g C m^-2^ yr^-1^, with total C and the birth and death components plotted on the same axes (death as a negative).
- Necromass trajectories with an explicit "vulnerable to loss" fraction (i.e., dead biomass that will burn in the next fire).
- How many trees are dying, and how does mortality vary with density and age?
- seems like there migth be a lot of recruitment in the mature stands..

## Demographics dataframe
- going to be comparing t1 to the last time point measured
- want a biomass measure of growth, recruitment, and death from the last time point by species and want this by age
- lots of big trees seem to recruit - hwo to handle this....I think get rid of for now but long-term or in the biomass terms of things
- for growht - should i be subtracting out the below dbh biomass? mabye...
- currently a few species not included

```{r}
# Growth
growth <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    Year_before = lag(Year), 
    biomass_diff = (above_dbh_biomass - lag(above_dbh_biomass)),
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3, this will be so we can get all of the BA and biomass of the live trees by species
  filter(sample.pd == 3) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>%
# get rid of the crazy high stuff - this will also get rid of the NAs!
  filter(Annual_Growth < 2 & Annual_Growth > -0.5) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass = sum(biomass_diff)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Growth")


# Recruitment
# below dbh to subtract out...
dbh.25 <- tree.age.traj %>%
  filter(DBH == 2.5) %>%
  select(SPECIES, DBH) %>%
  unique() %>%
  mutate(DBH2 = DBH - 0.01, biomass = allom.dbh(SPECIES, DBH2)) %>%
  select(-DBH)

recruit <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3 that have an NA growth rate from period one
  filter(is.na(Annual_Growth)) %>%
  filter(sample.pd == 3) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  #mutate(biomass_above = above_dbh_biomass - biomass) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Recruitment")

# join growht and recruitment

trees <- rbind(growth, recruit)

# Death
tree.lookup <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  filter(sample.pd == 3)

# identify trees that show as dead in the final smapling period
death.trees <- tree.age.traj %>%
  filter(DA == 0 & sample.pd != 1)%>%
  select(UniqueID)%>%
  distinct()

death <- tree.age.traj %>%
  filter(sample.pd !=3) %>%
  filter(UniqueID %in% death.trees$UniqueID) %>%
  filter(DA == 1) %>%
  group_by(UniqueID) %>%
  slice_max(order_by = sample.pd, n = 1)%>%
  ungroup()%>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    Year_diff = Year - lag(Year)
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  #filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area)*-1, 
   n_trees = n()*-1) %>%
  mutate(type = "Death")

# join them all and get rid of the few sites that burned during and FP5A and FP5D  
# need to add in the year component of htis figure!!
age_final <- ages %>%
  ungroup()%>%
  select(PLOT, Year, age)%>%
  group_by(PLOT)%>%
  slice_max(Year, n = 1) %>%
  mutate(age3 = 2026 - age)

age_diff <- ages %>%
  filter(sample.pd!= 2) %>%
  group_by(PLOT)%>%
  summarize(min_yr = min(Year), max_yr = max(Year)) %>%
  mutate(age_diff= max_yr-min_yr) %>%
  ungroup()%>%
  select(PLOT, age_diff)

# add in the age
trees <- rbind(growth, recruit, death) %>%
  left_join(age_final)%>%
  filter(!is.na(age3)) %>%
  group_by(SPECIES, age3) %>%
  mutate(
    plot_num = as.numeric(factor(PLOT)),
    x_plot = age3 + (plot_num - mean(plot_num)) * 0.2
  ) %>%
  ungroup() %>%
  left_join(age_diff)%>%
  mutate(age_diff = ifelse(age_diff==0 & age_category == "Young", 10, age_diff))%>%
  filter(Year %!in% c(2018, 2019)) %>%
  mutate(biomass_yr = biomass/age_diff) 
  

# Plot and biomass 
ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  fill = type))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~SPECIES, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# Plot and biomass - species
ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  fill = SPECIES))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~type + canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
# Plot and biomass - species - age category
ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  fill = SPECIES))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~type + age_category, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )


ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  color = type, fill = SPECIES))+
  geom_col() +
  theme_bw()+
  scale_color_manual(values = c("black", "red", "blue")) +
  scale_fill_manual(values = c("#E6AB02", "#66A61E", "darkgreen", "darkgoldenrod")) +
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~ canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

ggplot(trees, aes(x= reorder(PLOT, age3), n_trees/sample.area,  fill = SPECIES))+
  geom_col() +
  theme_bw()+
  ylab("No trees/m2")+
  facet_wrap(~type + canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
# Plot and biomass, adding in canopy
ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  fill = type))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~SPECIES + canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )


# plot and number of trees
ggplot(trees, aes(x= reorder(PLOT, age3), n_trees,  fill = type))+
  geom_col() +
  theme_bw()+
  facet_wrap(~SPECIES, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# more of a continuous scale
ggplot(trees, aes(
  x = x_plot,
  y = biomass,
  fill = type,
  group = PLOT
)) +
  geom_col(width = 0.35) +
  theme_bw() +
  facet_wrap(~SPECIES, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# each fraction individually with time
ggplot(trees, aes(age3, biomass_yr, color = log10(abs(n_trees/sample.area))))+
  geom_point()+
  theme_bw()+
  scale_color_viridis_c() +
  facet_wrap(~type + SPECIES, scales = "free")

```

## scaling up demographics to stand level
```{r}
glimpse(trees)
unique(trees$sample.pd)

stand.level <- trees %>%
  ungroup()%>%
  group_by(Network, PLOT, canopy, sample.area, age_category, age_diff, type, age3) %>%
  summarize(
    # total number of trees in each class
    tot_trees = sum(n_trees),
    # total biomass
    tot_bio_m2 = sum(biomass), tot_bio_m2_yr = sum(biomass_yr)
  ) %>%
  ungroup()%>%
  # create a new variable for trees/area
  mutate(tot_trees_m2 = tot_trees/sample.area)

glimpse(stand.level)

ggplot(stand.level, aes(x= reorder(PLOT, age3), tot_bio_m2_yr,  fill = type))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# each fraction individually with time
ggplot(stand.level, aes(age3, tot_bio_m2_yr, color = log10(abs(tot_trees_m2))))+
  geom_point()+
  theme_bw()+
  scale_color_viridis_c() +
  facet_wrap(~type + canopy, scales = "free")

```

## growth, mortality, and competition
- using stand basal area as metric of competition right now
```{r}
#glimpse(stand.2)

stand.2 <- stand %>%
  ungroup()%>%
  select(-c(sample.pd, age))
stand.trees <- left_join(trees, stand.2)
glimpse(stand.trees)


ggplot(stand.trees, aes(BA_1, biomass_yr, color = SPECIES, shape = canopy))+
  geom_point()+
  theme_bw()+
  facet_wrap(~type + age_category, scales = "free")

ggplot(stand.trees, aes(n_trees_1, biomass_yr, color = SPECIES, shape = canopy))+
  geom_point()+
  theme_bw()+
  facet_wrap(~type + age_category, scales = "free")

```

## death check

```{r}

death.2 <- tree.age.traj %>%
  filter(sample.pd !=3) %>%
  filter(UniqueID %in% death.trees$UniqueID) %>%
  filter(DA == 1) %>%
  group_by(UniqueID) %>%
  slice_max(order_by = sample.pd, n = 1)%>%
  ungroup()%>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    Year_diff = Year - lag(Year)
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>%
  group_by(PLOT, SPECIES, sample.pd, age_category, canopy, age) %>%
  summarize(mean_DBH = mean(DBH), sd = sd(DBH), n = n()) %>%
  mutate(age2 = 2026 - age)%>%
  mutate(status = "dead")

glimpse(death.2)

ggplot(death.2, aes(age2, mean_DBH))+
  geom_point()+
  theme_bw()+
  facet_wrap(~SPECIES)

live.comparison <- tree.age.traj %>%
  filter(sample.pd ==3) %>%
  filter(UniqueID %!in% death.trees$UniqueID) %>%
  filter(DA == 1) %>%
  group_by(UniqueID) %>%
  ungroup()%>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    Year_diff = Year - lag(Year)
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>%
  group_by(PLOT, SPECIES, sample.pd, age_category, canopy, age) %>%
  summarize(mean_DBH = mean(DBH), sd = sd(DBH), n = n()) %>%
  mutate(age2 = 2026 - age) %>%
  mutate(status = "live")

ggplot(live.comparison, aes(age2, mean_DBH))+
  geom_point()+
  theme_bw()+
  facet_wrap(~SPECIES)

dead.live <- rbind(death.2, live.comparison)

ggplot(dead.live, aes(age2, mean_DBH, color = status))+
  geom_point()+
  theme_bw()+
  facet_wrap(~SPECIES)

```


# leaving this 
- what happened to bdm1
- site with a lot of death (particularly with poptre) - what is coming in, is there a lot of loss

competition on growth and mortality - stand basal area and stand density index (N x(DBHmean/25.4)^1.605) where N is number of trees per hectare adn dbh mean is the quadratic mean of D/bh in cm


# 6. Stand composition

- Even mature sites appear to keep recruiting — are there interesting recruitment patterns within mature stands?
- Are there mature/intermediate sites with large proportions of black spruce death?
- Overyielding in mixed stands — is it driven by reduced self-thinning, pathogen resistance, or shifts in intraspecific competition?
- Stand grwoht – what matters: is the productivity of a few vs. many and how does that change voer time? Relationship between density and biomass and size and time
- for the sites that change trajectories - is it death or birth that drives those changes?  
- stand biomass over time

## stand biomass over time

```{r}
glimpse(tree.age.traj)
st.bio <- tree.age.traj %>%
  mutate(sample.area = Plot.Area.M2 * sampled_plots_period) %>%
  group_by(Network, PLOT, DA, sample.area, age, age_category, canopy, sample.pd) %>%
  summarize(stand_bio = sum(above_dbh_biomass, na.rm = T), n = n(), mean_DBH = mean(DBH)) %>%
  mutate(stand_bio_m2 = stand_bio/sample.area, n_m2 = n/sample.area, 
         age2 = 2026 - age,age_category = factor(
    age_category,
    levels = c("Young", "Intermediate", "Mature")
  )) %>%
  filter(DA ==1) %>%
  filter(!is.na(age_category))

# ggplot(data = subset(st.bio, st.bio$age_category == "Young"), aes(PLOT, stand_bio_m2, fill = as.factor(sample.pd)))+
#   geom_col(position = "dodge")+
#   theme_bw()+
#    theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
#   )+
#   facet_wrap(~canopy, scales = "free")
# 
# ggplot(data = subset(st.bio, st.bio$age_category == "Intermediate"), aes(PLOT, stand_bio_m2, fill = as.factor(sample.pd)))+
#   geom_col(position = "dodge")+
#   theme_bw()+
#    theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
#   )+
#   facet_wrap(~canopy, scales = "free_x")
# 
# ggplot(data = subset(st.bio, st.bio$age_category == "Mature"), aes(PLOT, stand_bio_m2, fill = as.factor(sample.pd)))+
#   geom_col(position = "dodge")+
#   theme_bw()+
#    theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
#   )+
#   facet_wrap(~canopy, scales = "free_x")


ggplot(data = st.bio, aes(x= reorder(PLOT, age2), stand_bio_m2, fill = as.factor(sample.pd)))+
  geom_col(position = "dodge")+
  theme_bw()+
   theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )+
  facet_wrap(~canopy+age_category, scales = "free_x")
# makng a stacked barplot that shows the "lost/dead trees again?)

ggplot(st.bio, aes(age2, stand_bio_m2, color = canopy))+
  geom_point() +
  theme_bw()

```


## changing trajectories - birth or death

```{r}
traj.change <- trajectories %>%
  ungroup()%>%
  select(PLOT, canopy) %>%
  unique() %>%
  group_by(PLOT)%>%
  filter(n()>1)

changes1 <- trajectories %>%
  filter(PLOT %in% traj.change$PLOT) %>%
  filter(sample.pd == 1)%>%
  ungroup()%>%
  select(PLOT, canopy)%>%
  rename(first_canopy = canopy) 

changes <- trees %>%
  filter(PLOT %in% traj.change$PLOT) %>%
  left_join(changes1)

ggplot(changes, aes(x= reorder(PLOT, age3), biomass_yr,  fill = SPECIES))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~type + canopy, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )


```

## Density vs. biomass vs. size

```{r}

```



# Data questions

## Have
- Tree inventory
- Stand ages

## Need
- JFSP seedling/sapling data; also CAFI?
- consistent measures like soil moisture, slope, elevation, etc. across the two networks
- new RSN data - follow up with Jamie now?

Particularly with decimation of aspen stands, what are the consequences for stand productivity if we lose all of these trees both from the timber production and carbon storage persepctive?


# Demographics 1,2, and 3

```{r}
# Growth
growth12 <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    Year_before = lag(Year), 
    biomass_diff = (above_dbh_biomass - lag(above_dbh_biomass)),
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3, this will be so we can get all of the BA and biomass of the live trees by species
  filter(sample.pd == 2) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>%
# get rid of the crazy high stuff - this will also get rid of the NAs!
  filter(Annual_Growth < 2 & Annual_Growth > -0.5) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass = sum(biomass_diff)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Growth")


# Recruitment
# below dbh to subtract out...
dbh.25 <- tree.age.traj %>%
  filter(DBH == 2.5) %>%
  select(SPECIES, DBH) %>%
  unique() %>%
  mutate(DBH2 = DBH - 0.01, biomass = allom.dbh(SPECIES, DBH2)) %>%
  select(-DBH)

recruit12 <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3 that have an NA growth rate from period one
  filter(is.na(Annual_Growth)) %>%
  filter(sample.pd == 2) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  #mutate(biomass_above = above_dbh_biomass - biomass) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Recruitment")

# Death
tree.lookup <- tree.age.traj %>%
  filter(sample.pd == 1 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  filter(sample.pd == 2)

# identify trees that show as dead in the final smapling period
death.trees <- tree.age.traj %>%
  filter(DA == 0 & sample.pd == 2)%>%
  select(UniqueID)%>%
  distinct()

death12 <- tree.age.traj %>%
  filter(sample.pd ==1) %>%
  filter(UniqueID %in% death.trees$UniqueID) %>%
  filter(DA == 1) %>%
  group_by(UniqueID) %>%
  slice_max(order_by = sample.pd, n = 1)%>%
  ungroup()%>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    Year_diff = Year - lag(Year)
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  #filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area)*-1, 
   n_trees = n()*-1) %>%
  mutate(type = "Death")

# join them all and get rid of the few sites that burned during and FP5A and FP5D  
# need to add in the year component of htis figure!!
age_final12 <- ages %>%
  ungroup()%>%
  select(PLOT, Year, age)%>%
  group_by(PLOT)%>%
  slice_max(Year, n = 1) %>%
  mutate(age3 = 2018 - age)

age_diff12 <- ages %>%
  filter(sample.pd!= 3) %>%
  group_by(PLOT)%>%
  summarize(min_yr = min(Year), max_yr = max(Year)) %>%
  mutate(age_diff= max_yr-min_yr) %>%
  ungroup()%>%
  select(PLOT, age_diff)

# add in the age
trees12 <- rbind(growth12, recruit12, death12) %>%
  left_join(age_final12)%>%
  filter(!is.na(age3)) %>%
  group_by(SPECIES, age3) %>%
  mutate(
    plot_num = as.numeric(factor(PLOT)),
    x_plot = age3 + (plot_num - mean(plot_num)) * 0.2
  ) %>%
  ungroup() %>%
  left_join(age_diff12)%>%
  mutate(age_diff = ifelse(age_diff==0 & age_category == "Young", 5, age_diff))%>%
  #filter(Year %!in% c(2018, 2019)) %>%
  mutate(biomass_yr = biomass/age_diff, period = "1to2") 


# Growth
growth23 <- tree.age.traj %>%
  filter(sample.pd == 3 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    Year_before = lag(Year), 
    biomass_diff = (above_dbh_biomass - lag(above_dbh_biomass)),
    biomass_change = (above_dbh_biomass - lag(above_dbh_biomass))/Year_diff, 
    biomass_RGR = (log(above_dbh_biomass) - log(lag(above_dbh_biomass)))/Year_diff, 
    age2 = Year - age
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3, this will be so we can get all of the BA and biomass of the live trees by species
  filter(sample.pd == 3) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>%
# get rid of the crazy high stuff - this will also get rid of the NAs!
  filter(Annual_Growth < 2 & Annual_Growth > -0.5) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass = sum(biomass_diff)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Growth")


# Recruitment
# below dbh to subtract out...
dbh.25 <- tree.age.traj %>%
  filter(DBH == 2.5) %>%
  select(SPECIES, DBH) %>%
  unique() %>%
  mutate(DBH2 = DBH - 0.01, biomass = allom.dbh(SPECIES, DBH2)) %>%
  select(-DBH)

recruit23 <- tree.age.traj %>%
  filter(sample.pd == 3 | sample.pd == 2) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(sample.pd, .by_group = TRUE) %>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    lag_DBH = lag(DBH),
    # BA in m2
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  ungroup() %>%
  # filter to just look at smaple round 3 that have an NA growth rate from period one
  filter(is.na(Annual_Growth)) %>%
  filter(sample.pd == 3) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  #mutate(biomass_above = above_dbh_biomass - biomass) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area), 
   n_trees = n()) %>%
  mutate(type = "Recruitment")

# join growht and recruitment

# Death
tree.lookup <- tree.age.traj %>%
  filter(sample.pd == 2 | sample.pd == 3) %>%
  filter(!is.na(DBH) & DA == 1) %>%
  filter(sample.pd == 3)

# identify trees that show as dead in the final smapling period
death.trees <- tree.age.traj %>%
  filter(DA == 0 & sample.pd == 3)%>%
  select(UniqueID)%>%
  distinct()

death23 <- tree.age.traj %>%
  filter(sample.pd ==2) %>%
  filter(UniqueID %in% death.trees$UniqueID) %>%
  filter(DA == 1) %>%
  group_by(UniqueID) %>%
  slice_max(order_by = sample.pd, n = 1)%>%
  ungroup()%>%
  mutate(
    # sample area
    sample.area = Plot.Area.M2 * sampled_plots_period, 
    Year_diff = Year - lag(Year)
  ) %>%
  mutate(site_yr = paste(PLOT, sample.rd, sep="_")) %>%
  # filter FP5A for now because not qutie right and FP5D
  filter(PLOT %!in% c("FP5A", "FP5D")) %>%
  filter(SPECIES %!in% c("POPBAL", "LARLAR", "PICEA")) %>% 
  # filter out species that are way too big!
  #filter(DBH < 9) %>%
  #left_join(dbh.25) %>%
  group_by(Network, PLOT, sample.pd, SPECIES, canopy, sample.area, age_category) %>%
  summarize(
   biomass= sum(above_dbh_biomass)/first(sample.area)*-1, 
   n_trees = n()*-1) %>%
  mutate(type = "Death")

# join them all and get rid of the few sites that burned during and FP5A and FP5D  
# need to add in the year component of htis figure!!
age_final23 <- ages %>%
  ungroup()%>%
  select(PLOT, Year, age)%>%
  group_by(PLOT)%>%
  slice_max(Year, n = 1) %>%
  mutate(age3 = 2026 - age)

age_diff23 <- ages %>%
  filter(sample.pd!= 2) %>%
  group_by(PLOT)%>%
  summarize(min_yr = min(Year), max_yr = max(Year)) %>%
  mutate(age_diff= max_yr-min_yr) %>%
  ungroup()%>%
  select(PLOT, age_diff)

# add in the age
trees23 <- rbind(growth23, recruit23, death23) %>%
  left_join(age_final23)%>%
  filter(!is.na(age3)) %>%
  group_by(SPECIES, age3) %>%
  mutate(
    plot_num = as.numeric(factor(PLOT)),
    x_plot = age3 + (plot_num - mean(plot_num)) * 0.2
  ) %>%
  ungroup() %>%
  left_join(age_diff23)%>%
  mutate(age_diff = ifelse(age_diff==0 & age_category == "Young", 5, age_diff))%>%
  #filter(Year %!in% c(2018, 2019)) %>%
  mutate(biomass_yr = biomass/age_diff, 
         period = "2to3") 


## trees
# filter these out because don't have all three sampling periods
three <- trees.traj %>%
  ungroup()%>%
  select(c(PLOT, sample.pd)) %>%
  distinct() %>%
  group_by(PLOT)%>%
  summarize(n=n()) %>%
  filter(n==3)


trees123 <- rbind(trees12, trees23) %>%
  filter(PLOT %in% three$PLOT)

# Plot and biomass 
ggplot(trees123, aes(x= reorder(PLOT, age3), biomass_yr,  fill = type))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~SPECIES + period, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# Plot and biomass - species
ggplot(trees123, aes(x= reorder(PLOT, age3), biomass_yr,  fill = SPECIES))+
  geom_col() +
  theme_bw()+
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~type + canopy+period, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
# Plot and biomass - species - age category
# ggplot(trees, aes(x= reorder(PLOT, age3), biomass_yr,  fill = SPECIES))+
#   geom_col() +
#   theme_bw()+
#   ylab("Biomass - g/m2/yr")+
#   facet_wrap(~type + age_category, scales = "free")+
#     theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
#   )


ggplot(trees123, aes(x= reorder(PLOT, age3), biomass_yr,  color = type, fill = SPECIES))+
  geom_col() +
  theme_bw()+
  scale_color_manual(values = c("black", "red", "blue")) +
  scale_fill_manual(values = c("#E6AB02", "#66A61E", "darkgreen", "darkgoldenrod")) +
  ylab("Biomass - g/m2/yr")+
  facet_wrap(~ canopy + period, scales = "free")+
    theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
  
```



