---
title: "RSN_Analysis_Plan"
author: "OLH"
date: "2026-03-04"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 20pt; } h2 {font-size: 18pt; } h3 {font-size: 16pt; } h4 {font-size: 14pt; } h5 {font-size: 12pt; } h6 {font-size: 10pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot,
)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```


# Data QA/QC Notes
*	Zach is updating code to correct for a few things identified from looking at annualized growth rates after initial data processing and going through the code
  + new growth, ID'd as dead and then comes back
  + trees with DBH < 2.5 cm - how to address those
*	Some minor QC in the data – to work with Mark on – i.e., trailing spaces, propagating errors with -9999 (to -9998, -9997), etc - should maybe address trailing spaces in the data with Zach
*	Sampling effort is not necessarily clearly documented in the inventory – does a plot not have data because it wasn’t collected (time constraints, etc.) or because there weren’t trees in the plot. I think with this just need to document the expected number of plots - Marks' indicated that patterns are because the trees do not exist in the data
*	2013-2014 – standing dead were measured at sites. These measurements should not persist, but they do have a DBH measurement with a 0 for DA. 

# RSN Analysis

# Data summaries

## Site Summary table

* Record length of each site, number of surveys, number of plots --- grouped by age
* No. of trees and proportion of different types, average size at T1 and average age at last survey

## Site Characteristics 

* Across site age, moisture, productivity, trajectory, etc.
* Some sites are lacking this data b/c in inventory but not RSN. Will try to understand best way to get this data from Mark. 
* Other site data that would be important to capture?

# Individual

* Growth over time; are trees growing faster than they did before?
* Trends in the data over time and how that vary among species/age class/landscape position
  + DBH Difference
  + Relative growth rate
  + Change in BA
  + Relativized BA change (%)
* It doesn't seem that this would be the best data to ask this question at the individual level? I think it's more these sorts of questions at the stand level? Is there an advantage of this type of data over tree ring data for this sort of question?

# Stand level

* Among age classes, landscape position, and stand structure 
  + Growth - DBH, BA, biomass
  + Mortality
  + Turnover
  + Density

# Questions

* DBH size (2.5 cm threshold) - for sites earlier that seem to have all seedlings measured too?


# Read in Data 

```{r}
# create data folders
dir.create(file.path("2026 Data"), showWarnings = F)
dir.create(file.path("2026 Data", 'RSN_Site'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data'), showWarnings = F)

# Identify desired files to download
site_info_file <- "RSN_Master_Site.csv"
qaqc_data <- "RSN_Inventory_QAQC_Final.csv"

# Download data files
# site info
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1PWk45RgTK47B-PX71QG6BazpJyAtidQ1")) %>% 
  dplyr::filter(name == site_info_file) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "RSN_Site", .$name))

# qaqc data
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1mT036D-xu_XPbJGupVoAKH2DK982oGxx")) %>% 
  dplyr::filter(name == qaqc_data) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "QAQC_Data", .$name))

# Read in site info and data files
site_info <- read.csv(file = file.path("2026 Data", 'RSN_Site', site_info_file))

data <-  read.csv(file = file.path("2026 Data", 'QAQC_Data', qaqc_data))

```

# Summary Tables

## Site and tree summary

```{r}

```


## Site characteristics summary
* important to note that a few sites are missing data

```{r}

```


