---
title: "RSN_Summary_0415"
author: "OLH"
date: "2026-04-15"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    collapsed: TRUE
    theme: spacelab
    fig_caption: TRUE
    number_sections: FALSE
---
# Setup
<style>h1 {font-size: 24pt; } h2 {font-size: 20pt; } h3 {font-size: 18pt; } h4 {font-size: 16pt; } h5 {font-size: 14pt; } h6 {font-size: 12pt; } </style> 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dpi = 600, progress = FALSE, cache = FALSE, fig.width=5,  fig.height=3.5, fig.align="center")

options(repos = c(CRAN = "http://cran.rstudio.com"))
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, googledrive, broom, supportR, lubridate,  gt, data.table, readxl, corrplot, car, emmeans,knitr, cowplot)

# Clear environment + collect garbage
rm(list = ls()); gc()

# create a negate function
`%!in%` = Negate(`%in%`)

```

# Downlaod data
- most recent QA/QC data
- site metadata - both the overall summary file adn the site-year that will help with the sampling effort

```{r}

# create data folders
dir.create(file.path("2026 Data"), showWarnings = F)
dir.create(file.path("2026 Data", 'RSN_Site'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data'), showWarnings = F)
dir.create(file.path("2026 Data", 'QAQC_Data', 'Updated'), showWarnings = F)

# Identify desired files to download
site_info_file <- "RSN_Master_Site.csv"
qaqc_data <- "RSN_Inventory_QAQC_Final_2026-04-09.csv"

# Download data files
# site info
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1PWk45RgTK47B-PX71QG6BazpJyAtidQ1")) %>% 
  dplyr::filter(name == site_info_file) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "RSN_Site", .$name))

# qaqc data
googledrive::drive_ls(googledrive::as_id("https://drive.google.com/drive/folders/1VR7WRVcF3behOWDoXtuFlN9wK0M24jOb")) %>% 
  dplyr::filter(name == qaqc_data) %>% 
  googledrive::drive_download(file = .$id, overwrite = T,
                              path = file.path("2026 Data", "QAQC_Data", "Updated", .$name))

# Read in site info and data files
site_info <- read.csv(file = file.path("2026 Data", 'RSN_Site', 'site_metadata_OLH.csv'))

data <-  read.csv(file = file.path("2026 Data", 'QAQC_Data','Updated', qaqc_data))

site_year <- read.csv(file = file.path("2026 Data", 'Data_checks', 'site_years.csv'))

```

# Post QAQC Processing
- remove FP1A
- Get rid of trees that are below 2.5 cm at DBH (Unless they shrink down)
- Get rid of trees that are only ever NA
- Remove one SALBEB obs

```{r}
data.2 <- data %>%
  # remove salbeb
  filter(SPECIES != "SALBEB") %>%
  # remove site FP1A
  filter(SITE != "FP1A")%>%
  # set up for removing small trees
  arrange(UniqueID, Year) %>% # Ensure data is in chronological order
  group_by(UniqueID) %>%
  # keep NA DBH, DBH > 2.5, or if the DBH exceeds 2.5, it will keep subsequent observatoins
  # this code above removes 4693 observations
  filter(is.na(DBH) | DBH >= 2.49 | cumany(DBH >= 2.49)) %>%
  ungroup()


# see which trees only have an NA - this is after the seedlings have been removed, so in many cases these had an intiial measurement that was NA and corresponded to somehting like 0.7cm
na.trees <- data.2 %>%
  group_by(UniqueID) %>%
  filter(all(is.na(DBH))) #%>%
  #distinct(UniqueID)

# remove the na trees only
data.25 <- data.2 %>%
  filter(UniqueID %!in% na.trees$UniqueID)


```

# Join the data
- have data.25 which is all of the trees in the analysis without metadata
- site_data corresponds to the site-year and site metadata combo
- data.25.site - is the tree inventory joined with all of the site level data

```{r}
# join the site-year data with the overall site-metadata
site_data <- left_join(site_year, site_info, by = "SITE")

data.25.site <- left_join(data.25, site_data, by = c("SITE", "Year"))

```

# Summary of site data

# Summary of inventory data

## Sampling effort
```{r}
mature <- ggplot(data = subset(site_year, site_year$age_class == "Mature"), aes(SITE, Year, color = as.factor(sample.rd))) +
  geom_point(size=2) +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Mature Stands")
mature

inter <- ggplot(data = subset(site_year, site_year$age_class == "Intermediate"), aes(SITE, Year,color = as.factor(sample.rd))) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Intermediate Stands")
inter

young <- ggplot(data = subset(site_year, site_year$age_class == "Young"), aes(SITE, Year,color = as.factor(sample.rd))) + 
  geom_point() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=8)) +
  ggtitle("Young Stands")
young

# graph of sites and age classes
site.yr.sum <- site_year %>%
  group_by(age_class_yr, sample.rd, sample.yr) %>%
  summarize(num=n())

ggplot(site.yr.sum, aes(sample.yr, num, group=age_class_yr, fill=age_class_yr))+
  geom_col(position="dodge", color="black")+
  theme_bw() +
  ylab("Number of Sites")+
  xlab("Sample Year") +
  theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )
```

## No of trees
```{r}
first_last <- data.25.site %>%
  group_by(SITE) %>%
  summarize(min = min(sample.rd), max = max(sample.rd)) %>%
  pivot_longer(cols = min:max, names_to = "min", values_to = "sample.rd") %>%
  mutate(census = ifelse(min == "min", "first", "last")) %>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_")) %>%
  select(site_yr, census)

data.summary <- data.25.site %>%
  group_by(SITE, sample.yr, sample.rd, SPECIES, age_class_yr) %>%
  summarize(n = n(), n_alive = sum(DA), n_dead = n - n_alive, mean_DBH = mean(DBH, na.rm = T), se_DBH = (sd(DBH, na.rm = T))/(sqrt(n()))) %>%
  ungroup() %>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_"))%>%
  filter(site_yr %in% first_last$site_yr) %>%  left_join(first_last)%>%
  group_by(census, SPECIES, age_class_yr) %>%
  summarize(n_trees = sum(n), n_dead = sum(n_dead), n_site = n())

wide_table <- data.summary %>%
  mutate(age_class_yr = factor(age_class_yr, levels = c("Mature", "Intermediate", "Young"))) %>%
  pivot_wider(
    names_from = c(census, age_class_yr),
    values_from = c(n_trees, n_dead, n_site)
  )
```

## DBH Summaries by age class and species
- I think also want these by first and last census?

```{r}
dbh_summary <- data.25.site %>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_"))%>%
  filter(site_yr %in% first_last$site_yr) %>%  left_join(first_last) %>%
  filter(!is.na(DBH)) %>%
  filter(SPECIES != "PICEA")

# Ensure age_class_yr is a factor for discrete coloring
dbh_summary$age_class_yr <-factor(dbh_summary$age_class_yr, levels = c("Mature", "Intermediate", "Young"))

first_dbh <- ggplot(data = subset(dbh_summary, dbh_summary$census == "first"), aes(x = DBH, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
  geom_density(alpha = 0.5) + 
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    title = "DBH Distribution during First Census",
    x = "DBH",
    y = "Density",
    fill = "Age Class"
  )
first_dbh

last_dbh <- ggplot(data = subset(dbh_summary, dbh_summary$census == "last"), aes(x = DBH, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
  geom_density(alpha = 0.5) + 
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    title = "DBH Distribution during Last Census",
    x = "DBH",
    y = "Density",
    fill = "Age Class"
  )
last_dbh
```

# Tree level metrics
- first calc the BA, RGR, etc at the tree level

```{r}
trees <- data.25 %>%
  filter(!is.na(DBH) & DA == 1) %>%
  group_by(UniqueID) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(
    lag_DBH = lag(DBH),
    # BA in m2
    BA = 0.00007854 * (DBH)^2,
    BA_diff = BA - lag(BA),
    Year_diff = Year - lag(Year),
    Annual_Growth = (DBH - lag_DBH) / Year_diff,
    RGR = (log(DBH) - log(lag(DBH)))/Year_diff,
    RBA = (log(BA) - log(lag(BA)))/Year_diff,
    Year_before = lag(Year)
  ) %>%
  #filter(!is.na(Annual_Growth)) %>%
  left_join(site_data, by = c("SITE", "Year")) %>%
  select(-c(TREE_raw:Flag_LastObs_SpeciesSpike)) %>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_"))
```

## some viz

```{r}
trees.2 <- trees %>%
  filter(site_yr %in% first_last$site_yr) %>%  left_join(first_last) 


last_RGR <- ggplot(data = subset(trees.2, trees.2$census == "last"), aes(x = RGR, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
  geom_density(alpha = 0.5) + 
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    title = "Relative growth Rate Last Census",
    x = "RGR",
    y = "Density",
    fill = "Age Class"
  )
last_RGR

last_RBA <- ggplot(data = subset(trees.2, trees.2$census == "last"), aes(x = RBA, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
  geom_density(alpha = 0.5) + 
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    title = "Relative BA growth Rate Last Census",
    x = "RBA",
    y = "Density",
    fill = "Age Class"
  )
last_RBA
```

# Site-level aggregation
```{r}
site_growth <- trees %>%
  ungroup() %>%
  filter(!is.na(Annual_Growth)) %>%
  # get rid of outliers 
  #filter(Annual_Growth < 4 & Annual_Growth > -4) %>%
  group_by(sample.rd, sample.yr, SPECIES, age_class_yr, SITE) %>%
  summarize(n_tree = n_distinct(UniqueID), n_plot = n_distinct(PLOT),
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()),
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_RBA = mean(RBA), median_RBA = median(RBA), se_RBA = sd(RBA)/sqrt(n())
            ) %>%
  arrange(SPECIES, sample.rd, sample.yr, age_class_yr) %>%
  # remove tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR")) %>%
  # get rid of obs with only 1 site
  # filter(n_plot >1) %>%
  # # requrie at least 5 trees
  filter(n_tree >4)


site_growth_ts <- trees %>%
  ungroup() %>%
  # make a new column for if a tree is large or small
  mutate(size = ifelse(DBH>10, "Large", "Small")) %>%
  filter(!is.na(Annual_Growth)) %>%
  group_by(sample.rd,sample.yr, SPECIES, age_class_yr, SITE, size) %>%
  summarize(n_tree = n_distinct(UniqueID), n_plot = n_distinct(PLOT),
            mean_dbh = mean(DBH), median_dbh = median(DBH), se_dbh = sd(DBH)/sqrt(n()), max_dbh = max(DBH),
            mean_AG = mean(Annual_Growth), median_AG = median(Annual_Growth), se_AG = sd(Annual_Growth)/sqrt(n()),
            mean_RGR = mean(RGR), median_RGR = median(RGR), se_RGR = sd(RGR)/sqrt(n()),
            mean_RBA = mean(RBA), median_RBA = median(RBA), se_RBA = sd(RBA)/sqrt(n())
            ) %>%
  arrange(SPECIES, sample.rd, sample.yr, age_class_yr, size) %>%
# remove tree species wiht limited coverage
  filter(SPECIES %!in% c("PICEA", "LARLAR")) %>%
  # get rid of obs with only 1 site
  # filter(n_plot >1) %>%
  # # requrie at least more than 4 trees (so min 5)
  filter(n_tree > 4)

```

# Aggregate at age class, species level, etc.

## By age class, species, and size
```{r}
ac_sample <-  site_growth%>%
  ungroup()%>%
  group_by(sample.yr, SPECIES, age_class_yr) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            ) %>%
  # must be in more than one site
  filter(n_site > 1)

ac_sample.size <- site_growth_ts %>%
    ungroup()%>%
  group_by(sample.yr, SPECIES, age_class_yr, size) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            )  #%>%
  #filter(n_site > 1)
```

## By species and size

```{r}

sp_sample <-  site_growth%>%
  ungroup()%>%
  group_by(sample.yr, SPECIES) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            ) %>%
  # must be in more than one site
  filter(n_site > 1)

sp_sample.size <- site_growth_ts %>%
    ungroup()%>%
  group_by(sample.yr, SPECIES, size) %>%
  summarize(n_tree2 = sum(n_tree), n_plot2 = sum(n_plot), n_site = n_distinct(SITE), 
            mean_dbh2 = mean(mean_dbh), se_dbh = sd(mean_dbh)/sqrt(n()), max_dbh = max(mean_dbh),
            mean_AG2 = mean(mean_AG), se_AG = sd(mean_AG)/sqrt(n()), 
            mean_RGR2 = mean(mean_RGR), median_RGR = median(mean_RGR), se_RGR = sd(mean_RGR)/sqrt(n()),
            mean_RBA2 = mean(mean_RBA), se_RBA = sd(mean_RBA)/sqrt(n())
            )  %>%
  filter(n_site > 1)

```

## Visualize growht over time

### Species

```{r}
ggplot(sp_sample, aes(sample.yr, mean_RGR2 *100
                      )) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) + (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) - (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  scale_color_manual(values = c(
    "#56B4E9"
  )) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL)+
        theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )



# look at only last two samplign periods

ggplot(data =  subset(site_growth, site_growth$sample.yr %in% c(2018, 2024)), aes(as.factor(sample.yr), mean_RGR *100)) +
   geom_boxplot(
                width = 0.5,
               outlier.shape = NA)+
    geom_jitter(size = 1.5,
  alpha = 0.7) +
  facet_wrap(~ SPECIES) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL) +
      theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )

```


### Species adn Size

```{r}
ggplot(sp_sample.size, aes(sample.yr, mean_RGR2 *100,
                      color = size)) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) + (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) - (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  scale_color_manual(values = c(
    "Small" = "#56B4E9",
    "Large" = "orange"
  )) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL)

ggplot(sp_sample.size, aes(sample.yr, mean_RBA2 *100,
                      color = size)) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RBA2*100) + (se_RBA*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RBA2*100) - (se_RBA*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  scale_color_manual(values = c(
    "Small" = "#56B4E9",
    "Large" = "orange"
  )) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "BA Growth (%)", color = NULL) +
    theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )

# look at only last two samplign periods

ggplot(data =  subset(site_growth_ts, site_growth_ts$sample.yr %in% c(2018, 2024)), aes(as.factor(sample.yr), mean_RGR *100,
                      color = size)) +
   geom_boxplot(position = position_dodge(width = 0.6),
                width = 0.5,
               outlier.shape = NA)+
    geom_jitter(position = position_jitterdodge(
    jitter.width = 0.1,
    dodge.width = 0.6
  ),
  size = 1.5,
  alpha = 0.7) +
  facet_wrap(~ SPECIES) +
  scale_color_manual(values = c(
    "Small" = "#56B4E9",
    "Large" = "orange"
  )) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL) +
      theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )


```


### Species and age class
```{r}
ggplot(ac_sample, aes(sample.yr, mean_RGR2 *100, color = age_class_yr
                      )) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) + (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) - (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL)+
        theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )



# look at only last two samplign periods

ggplot(data =  subset(site_growth, site_growth$sample.yr %in% c(2018, 2024)), aes(as.factor(sample.yr), mean_RGR *100, color =age_class_yr)) +
  geom_boxplot(position = position_dodge(width = 0.6),
                width = 0.5,
               outlier.shape = NA)+
    geom_jitter(position = position_jitterdodge(
    jitter.width = 0.1,
    dodge.width = 0.6
  ),
  size = 1.5,
  alpha = 0.7) +
  facet_wrap(~ SPECIES) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL) +
      theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )

```

### Species, size, & age class

```{r}
ggplot(data = subset(ac_sample.size, ac_sample.size$size == "Small"), aes(sample.yr, mean_RGR2 *100, color = age_class_yr
                      )) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) + (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) - (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL)+
        theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )+
  ggtitle("Small")

ggplot(data = subset(ac_sample.size, ac_sample.size$size == "Large"), aes(sample.yr, mean_RGR2 *100, color = age_class_yr
                      )) +
  geom_step(linewidth = 1, direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) + (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +
  geom_step(aes(y = (mean_RGR2*100) - (se_RGR*100)),
            linetype = "dashed",
            linewidth = 0.6,
            alpha = 0.7,
            direction = "hv") +

  geom_point(size = 1.5) +
  facet_wrap(~ SPECIES) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    legend.position = "bottom"
  ) +
  labs(x = "Inventory year", y = "Relative Growth Rate (%)", color = NULL)+
        theme(
    axis.title = element_text(size = 14), 
    axis.text = element_text(size = 14)
  )+
  ggtitle("Large")



```

# Site-level BA
- this is meant to look more at the sum vs. tree levle basaal area

```{r}
sl.ba <- trees %>%
  group_by(SITE, age_class_yr, sample.yr, sample.rd, SPECIES, Plot.Area.M2, sampled_plots_period) %>% 
  summarize(site_ba = sum(BA), area = mean(Plot.Area.M2) * mean(sampled_plots_period), site_ba_m2 = site_ba/area, site_ba_ha = site_ba_m2/0.0001)

sl.ba.sp <- sl.ba %>%
  ungroup()%>%
  group_by(SITE, sample.yr, sample.rd, age_class_yr) %>%
  summarize(tot_ba = sum(site_ba_ha))

sl.ba <- left_join(sl.ba, sl.ba.sp)
  
```


## Visualize

### Total
```{r}
sl.ba.tot.last <- sl.ba.sp%>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_"))%>%
  filter(site_yr %in% first_last$site_yr) %>%  left_join(first_last)

ggplot(sl.ba.tot.last, aes(tot_ba)) +
  geom_histogram(fill = "lightgrey", color = "black") + 
  theme_bw() +
  xlab("Basal Area (m2/ha)")+
  facet_wrap(~census) 

ggplot(data = sl.ba.tot.last, aes(x = tot_ba, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
  geom_density(alpha = 0.5) + 
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ census, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    x = "Basal Area (m2/ha)",
    y = "Density",
    fill = "Age Class"
  )

```

### By Species
```{r}
sl.ba.last <- sl.ba%>%
  mutate(site_yr = paste(SITE, sample.rd, sep="_"))%>%
  filter(site_yr %in% first_last$site_yr) %>%  left_join(first_last) %>%
  filter(census == "last") %>%
  filter(SPECIES != "PICEA")

ggplot(sl.ba.last, aes(tot_ba)) +
  geom_histogram(fill = "lightgrey", color = "black") + 
  theme_bw() +
  xlab("Basal Area (m2/ha)")+
  facet_wrap(~SPECIES) 

ggplot(data = sl.ba.last, aes(x = tot_ba, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
    geom_histogram(
    position = "identity",
    bins = 30,
    alpha = 0.4
  )+
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    x = "Basal Area (m2/ha)",
    y = "Count",
    fill = "Age Class"
  )

ggplot(data = sl.ba.last, aes(x = tot_ba, fill = age_class_yr)) +
  # alpha controls transparency (0 = clear, 1 = opaque)
    geom_density(
    alpha = 0.4
  )+
  # facet_wrap creates a separate plot for each species
  facet_wrap(~ SPECIES, scales = "free") + 
  # Clean up the look
  theme_minimal() +
  labs(
    x = "Basal Area (m2/ha)",
    y = "Density",
    fill = "Age Class"
  )
```


